---
title: 'The Political Origins of Education Decentralization'
author: "Thomas R. Vargas^[Postdoctoral Fellow, Department of Political Science, McMaster University, Hamilton, ON. Corresponding address: vargasct@mcmaster.ca. I thank Teri Caraway, Michelle Dion, John Freeman, Claire Le Barbenchon, Dave Lopez, Oanh Nguyen,  Oscar Pocasange, David Samuels, Ben Ross Schneider, and two anonymous reviewers for their comments on previous drafts of this paper. I am also grateful to seminar participants at the University of Minnesota and Duke University, and audiences at APSA 2019 and MPSA 2018. My thanks to the Department of Political Science at the University of  Minnesota for research support. Any remaining mistakes are my own.]"
abstract: \singlespacing What explains why some countries decentralize education while others shy away from such efforts? In the 1990s several Latin American countries reformed their education systems to give local communities more control over the management of public education. A prominent argument in the literature is that decentralization was pursued to improve the coverage and quality of education (Grindle 2004; Kaufman and Nelson 2004). Drawing on an in-depth case study of El Salvador with reference to Paraguay, and a statistical analysis of 93 presidential administrations covering 18 countries over 20 years, I show that education reform in Latin America was driven in part by political processes that had little to do with access in each country’s public education system. I argue that, depending on the configuration of political forces, education decentralization was pursued as part of an electoral strategy. Specifically, I show that incumbent political parties supported decentralizing reforms because they weakened teachers’ unions affiliated with opposition parties, thus depressing mobilization and votes for the opposition.
date: "This draft: March 7, 2021"
bibliography: setup/bib.bib
csl: setup/comparative-politics.csl
mainfont: Times New Roman
sansfont: Times New Roman
colorlinks: true
documentclass: article
geometry: margin=1in
fontsize: 12pt
linestretch: 1.5
link-citations: true
output:
  pdf_document: 
        includes:
          in_header: setup/header.tex
        number_sections: true
  html_document: default
  word_document: 
    reference_docx: setup/word-template.docx

---
\newpage

# Introduction

Public investment in primary education has huge payoffs in terms of long-term development outcomes. Many countries in the developing world, however, struggle to improve the quality and coverage of public education. One widely touted reform designed to improve such outcomes is *decentralization.* In this regard, scholars are presented with a puzzle: some countries around the world with low educational achievement and coverage rates decentralized primary and secondary education, while others with similar attributes did not. If---as prominent observers have suggested [@grindle_despite_2004; @kaufman_politics_2004]---governments pursue decentralization in order to improve access to education (a “technical” goal), the explanation for such variation remains a mystery. I argue, by contrast, that governments decentralized provision of education for explicitly political purposes, not primarily to improve access to education. Focusing on explaining variation across Latin America specifically, I argue that decentralization can strengthen the electoral position of incumbent political parties, depending on the political alignment of teachers’ unions.

Consider the cases of El Salvador and Paraguay. In 1991 El Salvador introduced a school-based management (SBM) program called *Educación con Participación de la Comunidad* (EDUCO) that would eventually devolve the governance and administration of most rural schools to local communities. Under EDUCO, rural communities oversaw infrastructure, collected school fees, and hired teachers—a truly extensive and radical rethinking of how to manage the provision of public education in the region. By contrast, Paraguay did not decentralize education despite the many similarities to El Salvador during its EDUCO years: it was ruled by a right-wing government, it had low primary education completion rates, and it was beset by high levels of poverty in rural areas (for other similarities, see table 1). While Paraguay experimented with some small changes to the educational curriculum, governance initiatives were muted.

The difference in decentralization decisions can be explained by looking at the partisan affiliations of teachers and their role in electoral strategies. Following the Salvadoran civil war (1980–1992), parties faced considerable uncertainty regarding their electoral prospects under democracy. The ruling right-wing party, *Alianza Republicana Nacionalista* (ARENA), feared losing power to a guerrilla-turned-legitimate political actor, the *Frente Farabundo Martí para la Liberación Nacional* (FMLN). To avert the electoral success of the left, ARENA sought to limit the ability of teachers’ unions to mobilize in favor of the FMLN, their political allies, by implementing education decentralization in politically sensitive rural areas. In Paraguay the absence of a substantive link between teachers’ unions and the opposition party meant there were no strong incentives to reform the governance of education, even though the education system’s performance (measured by completion rates, for instance) was as weak as El Salvador’s. In fact,  the largest teachers’ union was associated with the ruling right-wing party, the *Asociación Nacional Republicana* (the *Colorados*), making education decentralization politically unnecessary. These examples showcase the political logic that can help explain why some Latin American countries reformed the governance of education in the 1990s while others did not: incumbent political parties supported decentralizing reforms because they weakened teachers’ unions affiliated with opposition parties.

Much of the existing work that considers education decentralization focuses on how the process unfolds [@grindle_despite_2004;@kaufman_politics_2004;@corrales_politics_1999; @murillo_recovering_1999]. This study places less emphasis on the process through which decentralization is negotiated and implemented (i.e. outcomes) and asks a fundamental question others have not fully considered: why did countries advance decentralization reforms in the first place? I address this gap by bringing together the literature on the electoral advantages of political and fiscal decentralization for incumbents [@montero_decentralization_2004;@oneill_decentralization_2003;@oneill_decentralizing_2005] with scholarship exploring the partisan identities of teachers' unions [@larreguy_political_2017;@moe_comparative_2016; @murillo_recovering_1999], providing one of the first empirical accounts of the potential electoral benefits of education decentralization. I do not dispute the dominant paradigm: teachers can be powerful players in the politics of education, and their partisan attachments inform the strategies they follow in defending their vested interests [@moe_comparative_2016; @bruns_politics_2019].  But an emphasis on what teachers *do* (specifically, for instance, block reform) understates a central fact that structures democratic politics everywhere: parties want to win elections. Therefore, we should not assume that governments engage in politically costly changes to education systems exclusively because they want to improve them. Viewing education reform through this electoral lens can turn received wisdom on its head: the partisan attachments of teachers’ unions, broadly believed to empower them to resist change [@murillo_recovering_1999], can actually set them up to lose.

# What Motivates Governments to Advance Decentralization?

Recent contributions to the political science literature on decentralization focus on explaining *outcomes*. @falleti_decentralization_2010 argues that the order in which decentralization reforms are introduced shapes the resulting balance of power between levels of government. @niedzwiecki_social_2016 argues that decentralization arrangements and electoral incentives affect the quality of social policy implementation. Neither book wrestles, however, with the political motivations that lead governments to propose decentralization reforms. 

The choice to decentralize is not exogenous. Indeed, there is often a political bounty awaiting those governments that choose to empower lower levels of government [@oneill_decentralization_2003;@oneill_decentralizing_2005;@dickovick_municipalization_2007]. @oneill_decentralizing_2005 argues that incumbent political parties that fear losing power at the national level give up centralized authority to subnational units whenever they are more confident about their electoral chances at lower levels of government. In the context of India, @bohlken_democratization_2016 has shown that national governments introduce local government elections (a form of political decentralization) to control local intermediaries. In both of these arguments, parties engage in political decentralization with the expectation that their own political fortunes will improve in its aftermath. My point of departure is that politics can help explain why countries decentralize education. 

Explanations for education decentralization are lost in the larger decentralization scholarship because of its focus on wholesale (administrative, political, fiscal) decentralization in large federal countries. For example, in Colombia and Argentina, education decentralization was part of a larger decentralization package that was supposed to devolve more authority to subnational units of government. In such prominent cases it is difficult to distinguish the origins of education decentralization (a type of administrative reform) from the rest. But in small countries like El Salvador education decentralization was proposed and ultimately advanced alone, suggesting that this type of decentralization might have different underlying causes than the existing literature.

In the rest of this section, I look specifically to the education reform literature. Scholars in this tradition, like @kaufman_crucial_2004 and @grindle_despite_2004, suggest that the origins of education  decentralization spring from one of three causes: technical factors, ideational pressures, and diffusion. Each is insufficient to address the puzzle at the beginning of this paper. 

## The Technical Explanation

The technical determinants theory is the most prominent among scholars of education reform. It holds that some governments in the region decentralized because their public education systems were overwhelmed by a crisis of quality and coverage [@grindle_despite_2004;@kaufman_crucial_2004]. Through the 20th century, reformers argued, Latin American governments created bureaucratic behemoths that were too slow to meet these challenges. According to this line of thinking, the only viable solution to improving the quality and coverage of education was to break up concentrated authority. This would achieve gains in efficiency at the same time that it democratized decision-making by devolving decisions over education to local communities.

```{r, figure1, include=FALSE}
## Load packages
library(tidyverse)

## Load data
qog<-read_csv("data/qog_std_ts_jan19.csv") %>% # QoG dataset
  select("cname", "year", "ccodecow", "ccodealp", # select only relevant variables
                           "wdi_expedu", "bmr_dem", "p_polity2", "wdi_pop14",
                           "gle_cgdpc","gle_trade", "wdi_expeduge", "wdi_expedup",
                           "wdi_expedus", "ht_region",
                           "wdi_expedut", "wdi_expstup", "wdi_expstus", "wdi_expstut",
                           "bl_lpmf" , "wdi_popurb", "wdi_debt", "bl_asymf", "wdi_gerp",
                           "wdi_nerp") %>%
  rename("cowcode" = "ccodecow" ) 

## Merge data 

data<-read.csv("data/TRV_decentralization.csv") %>% # Decentralization data
  left_join(qog, by=c("cowcode", "year")) # Merge with QoG

## Plot performance data

data%>%
  select(full_name, year, wdi_gerp) %>%
  arrange(full_name)%>%
  subset(year>1977)

f1a<-data %>%
  ggplot(mapping=aes(x=year, y=wdi_gerp)) +
        geom_smooth(color="black", size=2) +
        geom_jitter(alpha = 0.2) +
        theme_bw() +
        xlab("Year") +
        ylab("Enrollment Rates") +
        coord_cartesian(xlim = c(1970,2010), ylim = c(0,120), expand = TRUE,
                        default = FALSE, clip = "on") +
        scale_x_continuous(breaks = seq(1970, 2010, by = 10)) 

f1b<-data %>%
  ggplot(mapping=aes(x=year, y=wdi_nerp)) +
        geom_smooth(color="black", size=2) +
        geom_jitter(alpha = 0.2) +
        theme_bw() +
        xlab("Year") +
        ylab("Completion Rates") +
        coord_cartesian(xlim = c(1970,2010),ylim = c(0,120), expand = TRUE,
                        default = FALSE, clip = "on") +
        scale_x_continuous(breaks = seq(1970, 2010, by = 10))

library(gridExtra)
```

```{r, plotfig1, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, fig.cap = "Enrollment and Completion Rates for School-Aged Children Latin America Since 1970", fig.width=5, fig.height=3, cache= F, label=fig1}
grid.arrange(f1a,f1b, ncol=2)
```

One problem with this argument is that the data available at the time governments began decentralizing did not unequivocally support the notion that all Latin American countries were experiencing a catastrophe in the education sector. Consider the share of children enrolled in primary school. Figure 1 plots enrollment and completion rates for Latin America on the left and right panels, respectively, alongside a data-smoothing trend line in black. The difference between enrollment and completion rates is that enrollment rates do not account for children repeating grades or starting late and thus can exceed 100%; completion rates correct for this issue. Both panels tell the same story: Over time, education systems have been able to improve the coverage of education. Moreover, the poorest performing countries have caught up quickly to regional trends. Whereas the lowest completion rates in 1970 hovered around 50%, by the year 2000 the lowest completion rates were well above 75%.

Still, it is plausible that regional trends in decentralization were driven by countries that, on average, trailed their peers. Several  authors have made the claim that voter demands and mass mobilization can compel governments to expand social policy [@garay_social_2016;@kosack_education_2012]. In our case, those countries with the lowest enrollment rates may have been under the most pressure to decentralize. This would be consistent with the notion that technical considerations propelled reform. As the cases of El Salvador and Paraguay illustrate, however, there are discrepancies even within countries that exhibited relatively weak coverage rates. In short, to account for regional variation in the decentralization of public education, we have to look beyond technical concerns regarding the coverage and quality of schools.  

## Ideational Theories 

The ideational theory of education reform attributes decentralization to neoliberal policy ideas that gained traction among center-right and right-wing parties. These parties, coming to power in the 1990s, were keen on introducing market-oriented strategies to the provision of social welfare. In the face of pervasive deficits, these governments were interested in keeping social expenditures sustainable and social policy programs cost-effective [@nelson_social_2011].  We would therefore expect right-wing governments to decentralize education, whereas center or left-wing governments would shy away from such efforts.  

The neoliberal logics of decentralization are easy to outline. For states burdened with debt, passing responsibility for the management and funding of education to subnational units would reduce central government spending. In theory, aggregate education expenditures would also fall, because the cost associated with collecting knowledge about prices and production processes is lowest at the local level [@shah_balance_1998]. Similarly, decentralization was thought to yield innovation in the provision of services as long as there was enough delegation of responsibility [@oates_fiscal_1972; @ford_rationale_2000; @tiebout_pure_1956].  For neoliberal advocates, decentralization offered the potential efficiencies and innovation associated with free markets.

Yet decentralization has not been consistently associated with a single political-economic logic [@montero_decentralization_2004]. During the time under analysis, both right- and left-wing governments proposed education decentralization programs. Take for instance, the example reviewed at the opening of this paper. While the case of El Salvador confirms the link between right-wing parties and decentralization, several null cases challenge it. For example, Paraguay, ruled by a right-wing party, did not engage in any decentralization efforts, whereas Honduras, ruled by a left-wing party, did.

```{r, setupfigure2, include=FALSE, cache= F}
## Load the Huber and Stephens Data
huber_stephens_political_data<-read_csv("data/hs_pol.csv") %>%
  mutate(left_parties = sl+scl+xl+xcl, 
         right_parties = scr+ sr + xcr + xr, # Create variables for Left and Right Parties in Congress
         right_in_the_executive = if_else(execpart == 4 | execpart == 5| execpart == 9 | execpart == 10,1, 0),
         country = as.character(contry), #rename country, set as factor
         country = replace(country, id=="CRI", "Costa Rica"), #replace labels for three countries so they merge
         country=replace(country, id=="ELV", "El Salvador"),
         country=replace(country, id=="DOM", "Dominican Republic")) %>%
  select(-federal) %>%
  mutate(country = as.factor(country),
         right_in_the_executive = as.factor(right_in_the_executive))

## Merge variables with the decentralization data

data <- data %>%
  rename("country" = "full_name") %>%
  left_join(huber_stephens_political_data,
            by=c("country", "year"))

## Plot graphs 
f2a<-data %>%
  filter(!is.na(right_in_the_executive)  & year>1979) %>% #Selects after 1979 those with parties
  ggplot(mapping=aes(x=right_in_the_executive, y=c_dec)) +
        geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
        theme_bw() +
        xlab("Right-Wing Executive") +
        ylab("Degree of Decentralization") +
        ggtitle("Executive Partisanship")

f2b<-data %>%
  filter(right_in_the_executive==1 & !is.na(execlhshare) & year>1979) %>% #filters only right parties
  ggplot(mapping= aes(x=execlhshare, y=c_dec)) +
        geom_smooth(color="black", size=2, method = "lm") +
        geom_jitter(alpha = 0.2) +
        theme_bw() +
        xlab("Executive Control of Congress") +
        ylab("Degree of Decentralization") +
        ggtitle("Right-Wing Parties")
```

```{r, plotfig2, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, fig.cap = "Decentralization and Right-Wing Neoliberals", fig.width=5, fig.height=3, cache= F}
grid.arrange(f2a,f2b, ncol=2)
```

Figure 2 examines this relationship more closely by exploring the association between right-wing executives (presidents) in Latin America and the degree of education decentralization. I borrow the orientation of the party of the executive from @huber_social_2012. The left panel of Figure 2 draws two violin plots showing the distribution of education decentralization for right- and left-wing administrations, coded 1 and 0 respectively. Violin plots provide all the information that boxplots do (for example, the median, lower, and top quartiles of the distribution), alongside the entire distribution of the data. The figure clearly shows that the distribution of decentralization is nearly identical across party ideological orientation.

Perhaps this reflects the inability of the executive to carry out neoliberal reforms due to the executive party’s minority status in the legislature. The right panel thus draws the degree of decentralization across the share of seats controlled by the executive in the lower house of the legislature. The data are exclusively for right-wing parties. Against conventional wisdom, the line has a *negative* slope, suggesting that more right-wing control across governments leads to *less* decentralization. In short, ideology is not a good predictor of decentralization. 

## Diffusion Explanations

Finally, the diffusion arguments suggest that governments began making certain types of changes to their education systems to accommodate both international financial institutions (IFIs) and changing global norms surrounding schooling. The early 1990s represented a unique time in international education. Donors and international organizations heavily promoted the idea that education was crucial to development. This is best exemplified by the World Conference on Education for All held in Thailand in 1990—financed by international organizations—which set five target education goals for 155 countries. One implied mechanism is that these organizations, including the World Bank (the Bank), pressured governments into adopting their own ideas about “what works” in education by funding certain education reform projects, or by attaching conditionalities to structural adjustment programs (SAPs). We would therefore expect that countries with active SAPs or sectoral loans with the World Bank would be more likely than others to decentralize education.

To be sure, international organizations helped shape the conversation around education reform in the 1990s [@edwards_jr_trajectory_2017;@mundy_world_2016]. A “global education agenda” was only possible because of the zeal with which agencies with massive convening power pursued the topic. Yet this literature overstates how emphatically agencies like the World Bank advocated for decentralization. For example, in working papers widely circulated around the Bank in the 1980s and the early 1990s, Bank staff warned that the decentralization of public services was no panacea, especially in poor countries with weak states [@rondinelli_decentralization_1983; @prudhomme_dangers_1995]. Moreover, the exact mechanisms through which international organizations and donors exerted pressure on reforming countries are underspecified. 

```{r, setupfigure3, include=FALSE, cache= F}
fig3a<-read_csv("data/imf_conditionality.csv") %>%
  mutate(condition_count=dBA1SP/dBA1TOT,
         hard_condition=dBA2SP/dBA2TOT,
         social_policy_sum= SP_IB+ SP_IB_W+ SP_PA+ SP_PA_W+ SP_QPC+
                            SP_QPC_W+ SP_SB + SP_SB_W + SP_SPC + SP_SPC_W,
         social_policy = if_else(social_policy_sum>0,1,0)) %>%
  ggplot(aes(x=year, y=dBA1TOT, fill=as.factor(social_policy)))+
    geom_bar(position = "fill", stat = "identity") +
  labs(x="Year", y="Share of Conditions",
       fill="Education Loans",
       title = "Global Loans") +
     theme_bw()+
  theme(legend.position="none")+
  scale_fill_manual(values=c("grey70", "grey20"))

fig3b<-read_csv("data/imf_conditionality.csv") %>%
  filter(cname=="Argentina"|cname=="Bolivia"|cname=="Brazil"|cname=="Chile"|cname=="Colombia"|cname=="Costa Rica"|cname=="Ecuador"|cname=="Dominican Republic"|cname=="El Salvador"|cname=="Guatemala"|cname=="Honduras"|cname=="Mexico"|cname=="Nicaragua"|cname=="Panama"|cname=="Paraguay"|cname=="Uruguay"|cname=="Venezuela, RB"|cname=="Peru") %>%
  mutate(condition_count=dBA1SP/dBA1TOT,
         hard_condition=dBA2SP/dBA2TOT,
         social_policy_sum= SP_IB+ SP_IB_W+ SP_PA+ SP_PA_W+ SP_QPC+
                            SP_QPC_W+ SP_SB + SP_SB_W + SP_SPC + SP_SPC_W,
         social_policy = if_else(social_policy_sum>0,1,0),
         redistribute_sum = POV_IB + POV_IB_W + POV_PA + POV_PA_W + POV_QPC +
           POV_QPC_W + POV_SB + POV_SB_W + POV_SPC + POV_SPC_W,
         red_policy = if_else(redistribute_sum>0,1,0))%>%
  ggplot(aes(x=year, y=dBA1TOT, fill=as.factor(social_policy)))+
    geom_bar(position = "fill", stat = "identity")+
  labs(x="Year", y="Share of Conditions",
       fill="Education Loans",
       title = "LAC Loans") +
     theme_bw()+
  theme(legend.position="none")+
  scale_fill_manual(values=c("grey70", "grey20"))
```

```{r, plotfig3, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, fig.cap = "Social Policy Conditions as Share of Total IMF Loan Conditions", fig.width=5, fig.height=3, cache= F}
fig3<-grid.arrange(fig3a,fig3b, ncol=2)
```

A close look at the data reveals that IFIs attached few conditionalities specifically demanding education decentralization in structural adjustment loans, nor in loans in support of social sector reform. Using data from @kentikelenis_imf_2016, hereafter KSK, I plot the share of conditionalities that reference the social sector (which includes education and health conditions) in Figure 3.  The left panel of Figure 3 examines social sector conditionalities for the entire global sample, showing that while these conditions were nearly non-existent in the early 1980s, they became exponentially more popular throughout the 1990s. The right panel shows conditionalities for the Latin American sample. The pattern is similar but not quite the same. The IMF seems to have placed exceedingly few education conditions on the social sector until the turn of the century, years after various decentralization initiatives had been completed. In short, while Latin American countries were big consumers of IFI loans, there is no clear evidence that the loans themselves required education sector reform. Indeed, during the period under analysis, both El Salvador and Paraguay held education sector reforms with the World Bank. While international organizations were important in shaping a specific environment in which decentralization was seen as a desirable or positive reform option, and they bankrolled various initiatives, this was not the principal *driver* of reform.

# The Argument

I argue that incumbent parties implement education decentralization programs as an electoral strategy. A first insight is that the costs of education governance decentralization fall disproportionally on teachers. Teachers have organizational and mobilizational advantages over most public employees. Not only do teachers typically outnumber every other public employee group by a fair margin[@bruns_great_2015], but they are also typically dispersed across all areas of a country and are deeply integrated into local communities [@moe_comparative_2016]. These advantages are compounded by centralized bargaining because teachers can credibly threaten to paralyze a country when a government fails to meet their demands. When education systems shift to decentralized governance, however, these advantages disappear because a single teachers’ union now must contend with multiple employers. More specifically, as a result of decentralization, teachers lose cohesion—because teachers face differing hiring and incentive structures across a national territory—and therefore negotiating leverage, at least with the central government [@olson_logic_1971]. In turn, teachers expect to face lower wages, less access to jobs, and worsening working conditions [@hanson_educational_1997; @murillo_recovering_1999]. For these reasons, teachers’ unions tend to strongly oppose decentralization [@grindle_despite_2004; @kaufman_crucial_2004], sometimes to the point of violence [@bruns_managing_2016]. 

A second insight is that in Latin America, as elsewhere in the developing world, parties rely on brokers both to collect information on voters and deliver votes [@stokes_brokers_2013; @szwarcberg_uncertainty_2012; @lowden_education_2004; @eaton_teachers_2014]. While scholars have previously pointed out how state employees can play the role of brokers [@weitzshapiro_what_2012; @oliveros_making_2016; @mares_core_2018], teachers are uniquely positioned to do so because of their distribution, numbers, and position of authority. Anecdotally, the notion that teachers’ unions have deep and steady partisan affiliations and deliver votes for their parties is common across the region.  @lowden_education_2004 reports that in Colombia it was typical for teachers to be hired *en masse* around election time. In Mexico, many teachers affiliated with the Mexican National Educational Workers' Union (SNTE) similarly owed their jobs to PRI leaders, who weaved a web of patronage-based practices regarding the hiring and management of teachers [@chambers-ju_teachers_2017]. Furthermore,  @larreguy_political_2017 have shown empirically that the PRI’s electoral victories were secured in part by the mobilizing efforts of the SNTE and that these attachments persist. Indeed, across much of the developing world, there is mounting evidence that teachers act as political brokers and that their unions are caught up in complex webs of clientelism and patronage [@sandholtz_voters_2019;@beteille_absenteeism_2009;@chattharakul_thai_2010;@pierskalla_personnel_2019;@eaton_teachers_2014]. Depending on the configuration of forces, certain parties’ reliance on unions for votes situates teachers’ unions as an arm of either the government or the opposition; therefore, they emerge as legitimate political targets of either group. If teachers’ unions’ partisan affiliations are relatively slow moving, then incumbents not aligned with teachers have an incentive to disrupt teacher organization. 

A final insight is that while teachers' unions are often successful in blocking reforms [@corrales_politics_1999;@murillo_recovering_1999;@diaz_rios_when_2016], circumstances surrounding changes to education governance made it difficult for unions to fully prevent them.  There are two reasons for this. First, incumbents were more successful than unions in leveraging what became a valance issue domestically and internationally: the improvement of public education. Once they mobilized against one decentralization proposal that was grounded in this rhetoric, unions were on weaker footing to oppose the next decentralization initiative. Teachers were thus forced to choose between full decentralization and compromising in exchange for policy concessions in other domains.  Second, democratic transitions across the region limited the intensity with which some unions advanced their interests. In some Latin America countries, fear of a regime reversal lead leftist elements to hold back on using all their pressure tools [@huber_social_2012;@caraway_working_2015]. In sum, despite being fully aware of the possible political costs of education decentralization, unions and their partisan allies could not prevent the strong winds of decentralization reform.

Previous work by Kathleen @oneill_decentralizing_2005 and others 
[@dickovick_municipalization_2007; @gonzalez_political_2008] has shown that incumbent parties undertake political and fiscal decentralization (e.g., strengthening subnational political units’ ability to raise taxes) when their support at the national level is waning. The logic of my argument builds on these contributions: in the face of competitive elections political incumbents can decentralize education as an electoral strategy. But unlike previous contributions, here I show how decentralization advantages incumbents by weakening *opponents* by targeting the organizing capability of teachers’ unions. By extension, where teachers and opposition parties are closely linked, decentralization limits the ability of unions to mobilize for their partisan allies. This electoral incentive can explain why some countries advanced education decentralization initiatives in the 1990s while others did not.

# Qualitative Evidence: El Salvador and Paraguay

To test this theory, I draw on a case study of El Salvador, described briefly at the opening of this paper, with reference to the contrasting case of Paraguay. Case selection follows a most-similar-systems design: despite their structural and political similarities, these countries differ on outcomes (see table 1). El Salvador pursued a broad-scale decentralization program that deeply transformed patterns of authority across public education, whereas Paraguay engaged in various education sector reforms that never truly decentralized education. By holding constant structural characteristics commonly used to explain why countries engage in decentralization, I show that these are not sufficient to explain patterns of education governance reform. The argument developed above can explain this puzzle.

\input{tables/table1.tex}

This analysis draws on four weeks of fieldwork in El Salvador, including interviews and reviews of historical material, two trips to the World Bank archives in Washington, DC, to assess documents that contained the history of EDUCO in the context of World Bank loans, and secondary sources. An interview methods section can be consulted in section A, page 3 of the appendix. The case study covers events that happened nearly 30 years ago, placing straightforward limitations on the data—key actors may no longer be alive, or they may adjust their recollections of events to fit dominant narratives. To overcome these issues, wherever possible, I triangulate interview data with archival data and secondary materials. The case of Paraguay is constructed mostly from secondary sources and a review of primary documents.

## El Salvador: Education Reform Under Electoral Competition

From 1940 to 1992 the provision of public education in El Salvador was strongly centralized, with the Ministry of Public Education (MINED) controlling the establishment, funding, and administration of public schools [@aguilar_aviles_vistazo_1998;@segovia_flores_historia_2012]. The Salvadoran civil war (1979–1992), however, severely affected the provision of public education in rural areas because government forces targeted teachers, who they assumed to be supportive of the FMLN.^[Interviews ES1, ES3, ES6] As a former guerrilla member puts it: “The guerrilla [forces] destroyed the economic infrastructure while the [government] army destroyed the education infrastructure; that was the division of labor.”^[Interview ES14] Moreover, reports by the @world_bank_salvador:_1990 and @unesco_servicios_1990 pointed to low completion rates and high repetition rates in public primary education in a context of falling central government investment in education. Thus, the Salvadoran government faced two typical problems for developing countries: improving the quality and coverage of education in a context of falling revenue.

Changing political circumstances opened a window for education reform. In 1989 Salvadorans elected the (right-wing) ARENA leader Alfredo Cristiani with a mandate to end the ongoing civil war. In anticipation of the signing of the Peace Accords (in 1992), government leaders were eager to return a sense of normalcy to daily life. This included returning official education services to rural areas and improving the quality of services.^[Interview ES14] A critical goal of any education reform was to “gradually decentralize management responsibilities to the regional and local levels” [@world_bank_staff_1993 pp.21] enshrined in the administration’s Socioeconomic Development Plan for the years 1990–1994. The idea was that decentralization, by virtue of delegating authority to lower levels of government, would create efficiencies in the provision of education at the same time that it improved the quality of education since parents would have a bigger say in the administration of schools and supervision of teachers. In sum, decentralization efforts were couched in technical considerations.

Government efforts to decentralize education were spearheaded by the Minister of Education at the time, Cecilia Gallardo, a hardline conservative with strongly held neoliberal views [@edwards_jr_trajectory_2017]. Gallardo’s menu of policy options was constrained by her ideology and dominant ideas in international education policy circles, both favoring some sort of decentralization strategy. More pressing, however, were the emerging political realities. During the civil war, local communities organized locally run schools known as *escuelas populares* to restore their access to education in places abandoned by the government. These *escuelas populares* served the guerrillas' ideological and political objectives. Ideologically, schools emphasized a state of war against injustice and inequality. Madre Rosa, a teacher, recounts: “I consider myself a *guerrillera* [guerrilla fighter] because [by teaching] we are fighting against a system that oppresses the poor and has never allowed them access to school, not even to learn their ABCs” [@de_la_cruz_educacion_1983 pp.95]. With the civil war coming to an end, it became imperative to exclude leftist elements and their institutions from areas previously controlled by the guerrillas.

The political salience of education reform was highlighted by the upcoming elections. Under the terms of the peace agreement, the FMLN would be allowed to participate as a political party in free elections in exchange for dismantling its military structure [@de_soto_implementation_1995]. In the run-up to the 1994 elections, there were thus two major political parties: ARENA and the FMLN. ARENA believed the elections were going to be competitive and viewed the FMLN as a real political contender.^[The FMLN did not nominate its own presidential candidate, but it did join a coalition of parties to present a nominee. It did, however, run its own candidates for the legislature.]  As reported by @lehoucq_election_1995, “[u]ncertainty about the electoral strength of the Left unnerved many in government, the military and their conservative supporters” (181). These groups acted accordingly. ARENA-affiliated death squads implemented a low-level campaign of violence against FMLN leaders. Prior to the elections, six leaders of the FMLN political coalition were murdered by right-wing death squads. ARENA also took to the airwaves, warning voters that an FMLN victory would threaten the peace and imply a “return to the past” [@stahler-sholk_salvadors_1994]. In addition, there were widespread claims of irregularities in the election process [@lehoucq_election_1995].

This political context affected the strategies pursued by the government to reform public education. Extending public schooling to rural areas could have been achieved by using the existing education governance system, but education leaders thought it would be too slow.^[Interview ES11] Most importantly, public school teachers and their organizations were viewed with suspicion. Most public school teachers belonged to the *Asociación Nacional de Educadores Salvadoreños, 21 de junio* (ANDES), teachers' union. ANDES was viewed as a tool of the radical left with deep partisan attachment to the FMLN opposition.  To some, the history of ANDES and the FMLN is one and the same. In 1970 the Secretary General of ANDES, Mélida Anaya Montes, co-founded the *Fuerzas Populares de Liberación* (FPL), one of the five component groups of the FMLN. Thus, in my interviews it was common for individuals to lump both organizations together. As a prominent right-wing politician told me, “The guerrilla is *born* in the teachers’ unions. The first guerrilla fighters of El Salvador are teachers”.^[Interview ES10]  While most ANDES-affiliated teachers were not directly involved in the conflict, there is some evidence that their organization provided support to the FMLN through union dues. In a cable from 1986, the CIA reports that Miguel Castellanos, a high-ranking FMLN defecting officer, stated that for all intents and purposes FMLN controlled ANDES, and that about half of all economic support it received was funneled to the guerrillas.^[``El Salvadors’ Insurgents: Resurrecting an Urban Political Strategy” Copy 181, from September 1986 held in the CIA Archives. Sanitized Copy Approved for Release 2011/12/30 accessed on January 21, 2019 available: [here][link1].]

[link1]: https://www.cia.gov/library/readingroom/docs/CIA-RDP04T00794R000100260001-0.pdf

The organizational closeness between the FMLN and ANDES makes clear that the latter's mobilizational advantages were key to the opposition's electoral strategy. @sprenkels_after_2018, who entered Chalatenango in 1992, recalls: "All the communities participated in party organizing, practically down to the last family." In places where the guerrillas did not have a visible presence, they exerted control through other agents; for example, “the ANDES teachers' union had departmental (subnational) organizational networks in Chalatenango within which the FPL  maintained strong contacts” (p.88). In the run-up to the 1994 elections, teachers formed a large part of the FMLN's party infrastructure. Most importantly, education leaders were painfully aware of this union-opposition bond and sought to disrupt it. In a 1991 interview Cecilia Gallardo, then Minister of Education, complained about ANDES’ political activity: “What ANDES should do is put on the FMLN’s t-shirt, to which they have always belonged by the way, and go campaigning,”^[From a newspaper article in 1991. “Una crisis de libros y lectores: Licenciada de Cano," La Prensa Gráfica (February 11, 1994).]  implying that the teachers’ union was merely an extension of the FMLN, and that their organizational advantages were used not only to secure the interests of their affiliates but also to support their partisan allies.

The dilemma facing the government, then, was how to place the *escuelas populares* and newly built schools under state control without granting the FMLN a further electoral advantage in rural areas by placing their allies and brokers—ANDES—in these communities. ARENA and Cecilia Gallardo found balance in EDUCO, a decentralization initiative under which schools were managed by elected parent councils, or ACEs,^[Acronym for *Asociaciones Comunales para la Educación*] but the central government maintained authority over naming principals, funding schools, and setting the curriculum.

Two features of the program reveal its political logic. First, EDUCO was most heavily targeted toward rural areas, and specifically, FMLN strongholds. As noted by a politician: “The EDUCO model served as a way to transition the guerrilla territory [from the FMLN to the government].”^[Interview ES11] In the period between 1991 and 1993, EDUCO quadrupled in size by targeting communities precisely in FMLN-controlled areas. On the eve of the 1994 elections, EDUCO covered roughly 2,300 teachers and 74,000 students (about 17% of all primary school children), almost all of them in rural areas [@ornelas_education_2019].

Second, EDUCO was built to exclude ANDES-affiliated teachers from teaching in rural areas. Under EDUCO teachers were hired on yearly contracts and were initially paid less than teachers in the official system.^[Eventually the World Bank, seeing EDUCO teachers jump ship whenever they could, advocated for a raise in the EDUCO pay schedule. This was after the 1994 elections. One of the unanticipated ramifications of the creation of parallel teaching tracks was that every EDUCO teacher wanted to enter the non-contracting option (Interviews ES1, ES6, ES12).] Teaching in EDUCO schools was therefore not appealing to tenured teachers affiliated with ANDES.^[Interviews ES5, ES6, ES12] At the same time, EDUCO teachers were required to meet a minimum set of teaching qualifications—i.e., obtaining some sort of degree. This had the effect of excluding instructors who had been teachers in *escuelas populares*.^[To allay the concerns of parents who had warmed up to teachers in *escuelas populares*, the Ministry of Education offered these teachers a path to formalization (Interview ES12). Instructors who had taught in *escuelas populares* would be eligible to teach under EDUCO if they underwent special trainings (Interview ES11). In practice, however, this had the effect of limiting the number of teachers in *escuelas populares* who would become EDUCO teachers, since it entailed leaving their posts, unpaid, for a significant amount of time. While it might be the case that 80 to 90% of teachers accepted the Ministry’s offer, as suggested by a former Ministry official (Interview ES11), the immediate job openings clearly presented a political opportunity for ARENA.] Finally, available teaching jobs were offered to ARENA sympathizers. In theory, the hiring of teachers was run through the ACEs. In practice, however, ACEs would often rubber-stamp the decisions made by government-appointed principals.^[Interviews ES4, ES7, ES8] The relationship between school principals and rural parents (some unable to read) was unequal, and principals thus maintained substantial authority in their communities. In one of my interviews a former EDUCO teacher admitted that he had been hired on the recommendation of a local ARENA party leader who had contacted the school principal.^[Interviews ES9] While it is impossible to know how common this practice was, this anecdote showcases that at different levels of government, the party’s commitment to EDUCO as a tool to improve educational outcomes was limited.

A final question remains: why would ANDES and the FMLN idly stand by as their political mobilization apparatus was dismantled? ANDES and its partisan allies were unusually weak in the aftermath of transition for two reasons. First, Gallardo was clever in designing the reform, both by developing a program (EDUCO) with roots in rebel organizations and by leveraging the support of the international community and prominent local actors.^[Interviews ES1, ES2, ES3] The fact that EDUCO schools were based on *escuelas populares* was meant to signal the program as a compromise with left-wing actors, despite the fact that it fully gutted *escuelas populares* and replaced them with a scheme favorable to the government. Moreover, the support of international organizations, including the World Bank, on a topic that all agreed was important—the improvement of education—placed ANDES on the defensive.

Second, most political actors viewed the transition period as tenuous, and therefore acted cautiously. Interviewees on the left and the right confirmed that the general sense among the population was one of exhaustion with social conflict.^[Interviews ES11, ES14] Thus, while ANDES mobilized large protests on a few occasions against education decentralization reforms, including a more severe decentralization program modeled after the Chilean experience,^[Interviews ES1, ES3] it eventually found itself bargaining with the government over the program. What did ANDES leaders extract from the government? The evidence is unclear, but some of my interviewees suggested that Gallardo promised ANDES leaders that most urban teachers (their electoral base) would remain excluded from the EDUCO system.^[Interviews ES5, ES11] Division within the FMLN was also significant. Some officers believed that an FMLN win would reverse democratization, splintering the organization.^[Interviews ES14] Thus, the FMLN was in a weak position to defend ANDES. In short, the specific dynamics of the transition and post-conflict period meant two things: 1) incumbents were relentless in their efforts to decentralize education while 2) ANDES and the FMLN were in no position to fully prevent it. The result was education governance decentralization.

The Salvadoran case study shows the political-electoral origins of education reform. The presence of a opposition-union link incentivized ARENA to advance a very specific decentralization program meant to weaken the organizational capabilities of teachers, and therefore the opposition. EDUCO cannot be explained simply as a strategy to expand public education to war-torn regions; the logic of its expansion and its operation in practice reveals a very carefully targeted program. An unresolved question is whether education leaders, aside from their political motivations, actually believed that the program would improve the coverage and quality of education. My argument is entirely consistent with this possibility: even if they thought EDUCO would improve education outcomes, it is improbable that they would not have seen an electoral benefit to the excluding of ANDES-affiliated teachers from the countryside. Indeed, in the absence of such an incentive, reform would not have been advanced.

## Paraguay: Limited Reform in the Absence of Electoral Incentives

The history of public education governance in Paraguay is like that of El Salvador. Starting in 1933 the national government made a concerted effort to control education directly, generating a fairly centralized education system which remained untouched for most of the century. By the end of the 1980s this education system had delivered poor results, especially in comparison with other Latin American countries. Reports commissioned by the @world_bank_paraguay_2002 pointed to low completion and achievement levels among students, particularly those from rural communities and from poor backgrounds. In 1990 primary completion rates sat at 65%, roughly on par with El Salvador, and were lower still for the rural student population which accounted for half of all students [@world_bank_world_2018]. In sum, there was a techincal case to be made for the improvement of public education. 

The overarching concern of the conservative party, the *Colorados*, however, was to remain in power under democracy, not education reform. General Stroessner was toppled in 1989 by his chosen successor General Andrés Rodríguez, who subsequently initiated a transition to democracy. Democracy brought new challenges for incumbents. The first sign of trouble emerged in 1991 when Carlos Filizzola of the *Asunción Para Todos* party was elected mayor^[Technically, Filizzola was elected *intendente municipal*.] of Asunción despite trailing the *Colorado* candidate Juan Manuel Morales in official polling.^["Paraguay celebra sus primeras elecciones municipales directas." El País, 26 of May, 1991, accessed on March 15, 2020 available: [here][link2].] Filizzola's victory offered a warning: in the post-transition period, the Colorados could not assume they would triumph at the ballot box [@velazquez_seiferheld_relaciones_2018]. In the run-up to 1993 this was confirmed by official polling, which placed the *Colorado* candidate, Juan Carlos Wasmosy, in an embarrassing third place. 

[link2]: https://elpais.com/diario/1991/05/27/internacional/675295206_850215.html

Counter to the expectations of the technical and ideational arguments, education reform under the right-wing *Colorados* was slow-moving. Recognizing that *something* had to be done, incumbents took a hands-off approach. In 1990 President Rodríguez established an education reform council, *Consejo Asesor de la Reforma Educativa* (CARE), and charged it with charting the path forward on education. To the relief of reformers, council members were well-respected independent experts who exhibited relative autonomy from the government in their deliberations [@nunez_reforma_2002]. Nonetheless, many of the ideas CARE considered were influenced by what was on offer by various international organizations, including the World Bank, USAID, and the Harvard Institute for International Development [@rodolfo_alisis_2014]. This meant that CARE considered decentralization a pillar of education reform [@rivarola_reforma_2000]. Indeed, CARE's first proposal, in 1992, centered reform around three objectives: 1) administrative changes in the Ministry of Education (MEC), 2) the selection of teachers, and 3) decentralization.^["Reforma Educativa, Compromisode Todos" 1992.] Much like in El Salvador, pressures pushing for decentralization were prominent among reformers. 

Unlike in El Salvador, these ideas were not being promoted by the government. The reason is that teachers were not allied with the opposition; in fact, for some teachers the reverse was true. During the Stroessner era, public sector jobs were tightly controlled by the dictatorship. To be eligible for a teaching position, for instance, individuals  needed to be *Colorado* party members and required a written recommendation from a mid-tier party or military leader [@rivarola_reforma_2000]. In the post-transition period, the largest teachers' union, the *Federación de Educadores del Paraguay* (FEP), with roots in dictatorship, was aligned with the *Colorados* [@lopez_sindicatos_2009; @becker_sindicatos_2009]. Predictably, the FEP mobilization was an important part of the *Colorado* electoral strategy. In the run-up to the 1993 elections, teachers were mobilized in rural areas to get out the vote in favor of the *Colorados*, leading to a high-profile clash with the Minister of Education, Raúl Sapena Brugada, who resigned rather than oversee a partisan education system [@nickson_wasmosy_1997; @harding_paraguay_1993]. 

In this context, promoting education decentralization would have weakened the incumbent's partisan allies. Theoretically, it could nevertheless advance *Colorado* interests by weakening non-FEP teachers. Other teachers' unions were active around this time, including the *Organización de Trabajadores de la Educación del Paraguay-Sindicato Nacional* (OTEP) and the *Unión Nacional de Educadores* (UNE) [@becker_sindicatos_2009]. These organizations had a contentious relationship with the incumbents, deploying strikes to bargain over the improvement of teacher working conditions. OTEP-led protests in 1990 and 1993, culminated in significant wins for teachers, including a flat wage increase of 30% [@nunez_reforma_2002]. If education decentralization hurts the organization of teachers' unions, pursuing decentralization as CARE suggested could have also brought about a more docile labor movement, to the benefit of the right-wing government.

Yet since OTEP and UNE were not aligned with the opposition, the *Colorados* would have gained no substantial electoral payoff from antagonizing this potent, organized force. Efforts to decentralize education therefore fizzled, despite a larger push to grant subnational units (*departamentos* and *municipios*) political and fiscal authority in the 1992 constitution. For example, the 1993 *Ley Orgánica Municipal* provided that municipalities *could* run their own schools if they so wished, but apportioned no national funds for such activity. Indeed, with the exceptions of Asunción y San Lorenzo there are no municipal schools in Paraguay [@gaete_estudio_2012]. At the department level, the 1998 *Ley General de Educación* mandated that the MEC establish *Consejos Departamentales de Educación* (CDE) to coordinate efforts with the *departamentos*. But CDEs are at best weak, advisory bodies to the MEC; they do not hire teachers, set education policy, nor do they administer or manage schools [@letelier_s_education_2018]. 

In sum, by the end of the 1990s education in Paraguay can only be characterized as strongly centralized, if with some allowances for subnational units to convey concerns over education to national authorities. Looking back on the promise of the education reform movement, the perception among many reform advocates is that things did not change radically. In the evaluation of a teacher not affiliated with FEP: "they have engaged in curriculum reforms, that theoretically and pedagogically where OK...but everything pretty much remains as before" [@palacios_de_asta_descentralizacion_2002]. This uneven commitment to decentralization by the *Colorados*--who remained in power until 2008, and thus had plenty of further opportunities to pursue reform--can only be explained by the absence of a opposition-union link. 

# Quantitative Evidence: A Regression Analysis

Do the patterns that emerge in El Salvador and Paraguay hold across other contexts? I develop a statistical test of my central hypothesis and find that it holds up to cross-national analysis across Latin America.

## Research Design

My investigation centers around assessing the association between instances of education reform and the presence of a teachers’ union link to the opposition (hereafter, OUL for opposition-union link). My main linear regression takes the form:


$$y_{ij} = \alpha_{[j]} + \text{OUL}_{ij}\zeta + \mathrm{X}_{ij}B + \epsilon_{ij} $$


Where $y_{ij}$ is the decentralization outcome at the end of administration $i$ in country $j$, $\alpha$ is randomly varying intercept for country $j$, $\textbf{X}$ is a vector of covariates for administration $i$ in country $j$, $\epsilon$ is an idiosyncratic error, and OUL is an indicator for whether teachers are affiliated with the opposition during administration $i$ in country $j$. The association of OUL with decentralization is captured by the $\zeta$ term. I estimate a linear probability model because it makes the interpretation of the coefficients easier, but the results remain largely unchanged when modeled using a logit link. A discussion of modeling choices, including the use of a hierarchical model, and Bayesian methods, results from a logit model, and information on priors is available in section B, page 7 of the appendix. I highlight only here that I set weakly informative priors over the parameters and run four chains, collect 10,000 draws from the posterior, discard the first half as burn-in, and summarize the results in section 5.3. 

## Data

I construct an original dataset of education governance at the primary level covering 93 presidential administration across 18 Latin America countries over 20 years (1980-2000). I chose to focus on presidential administrations since the electoral politics story that is at the foundation of my argument operates at the level of national elections. It is presidents and cabinet members who decide when and how to decentralize (or recentralize) the governance of education. The time period encompasses all major developments in the recent history of education reform in Latin America (i.e., decentralization initiatives such as El Salvador’s EDUCO), as well as several key rival hypotheses (i.e., the debt crisis, neoliberalism). Given that the arguments rests on electoral politics, countries enter the dataset only if they are democracies by minimal definitions.

The dependent variable is an indicator for whether there was an education governance decentralization reform. I create a categorical variable that ranges from 0 to 1, increasing in intervals of 0.25, where 0 reflects a strongly centralized system and 1 a strongly decentralized system. This coding strategy allows me to differentiate between countries where decentralization reforms have been fully implemented and those where it has been partial. The coding scheme can be found in section B.4.1, page 9 of the appendix. For the purposes of the subsequent analysis, I compare whether there is a decentralizing change at the end of a presidential administration compared to when it came into power. I code as 1 if the reform decentralizes education governance. The data contain 15 instances of decentralization across 93 administrations. Regressions therefore predict the probability of a decentralizing reform within an administration.

The main independent variable is the partisan attachment of teachers. For each presidential administration I evaluate whether the opposition can count on the support of teachers’ unions in their electoral coalition. This generates a binary variable, which I call opposition-union link (OUL), that takes the value 1 when present and 0 otherwise. This entailed reviewing secondary sources and carefully examining the history of teachers’ unions in the region. I define the opposition as the principal party opposing the incumbent in the previous election. I make no distinction across teachers’ unions: if the opposition counts on the support of *any* teachers’ union, then I code them as 1. This coding strategy allows me to capture the idea that the opposition’s attachments to any kind of teachers’ union (small or large) provide an incentive for decentralization. My models thus provide a probable underestimate of the effects of partisan attachments of teachers on education governance reform.

In addition, I adjust for a series of covariates that capture the competing explanations at the outset of this paper. These include 1) the performance of the education sector measured by enrollment rates from UNESCO, 2) the ideological center of gravity, which I reweigh to capture the strength of right-wing parties in the legislature, from @huber_politics_2008,  and 3) the number of  conditions attached to social policy loans by the IMF from @kentikelenis_imf_2016. I measure the last two variables at the end of the first year of each presidential administration. Summary statistics are available in section B.4.2, page 10 of the appendix. 

The general expectation is that incumbent parties decentralize education when completion rates are low, when the right is in power, and as a consequence of international organizations placing pressure to reform social policy expenditures. Finally, to account for democratizing forces, I include a measure of democratization. 

## Results

```{r, regdata, include=FALSE}
## Bayesian Regressions

## Load Packages
library(tidyverse)
library(stargazer)
library(bayesplot)
library(rstanarm)

## Read in data
pol<-read.csv("data/pres_political_data.csv") %>%
  mutate(year = as.factor(election_year),
         rates = completion_rate/10,
         oul = teachers_with_opp)

mini<-pol %>%
  select(education_decentralization_reform, completion_rate,
                        right_center_of_gravity, imf_education_loans, teachers_with_opp, 
                        polity, oul)
stargazer(mini) # For appendix
```

```{r, eval=FALSE, include=FALSE}
## Let's try something slightly different

## Data from Duke's DALP

# b6, vote-buying
# In general, how much effort do politicians and
# parties in this country make to induce voters with
# preferential benefits to cast their votes for them?

# b7 intense_vote_buying
# Do politicians nowadays make the same, greater
# or lesser efforts to provide preferential benefits to
# individuals and small groups of voters than they
# did about ten (10) years ago?

dalp <- read_csv ("data/partylevel_20130907.csv") %>%
  mutate(rely_on_unions_for_exchange = b10_1,
         mobilization_sucess = b11,
         rely_on_unions_for_monitoring =c2_1, 
         oul= a8_1, 
         oul_weight = oul*partysize) %>%
    select(country, ccode, party, pengname, partysize,rely_on_unions_for_exchange, mobilization_sucess,
           rely_on_unions_for_monitoring, oul, oul_weight) %>%
  group_by(country) %>%
  summarise(oul = sum(oul_weight),
            share_roufe = sum(rely_on_unions_for_exchange)/n(), 
            share_roufm = sum(rely_on_unions_for_monitoring)/n())

## I will look at whether countries in which parties rely on unions are associated with higher levels of decentralization. 

global_decentralization<-read_csv("data/decentralization_data_world.csv") %>%
  left_join(dalp, by="country")

rai<-read_csv("data/rai.csv")%>% # Data from the Regional Authority Index
  filter(year>2000) %>%
  mutate(country=country_name)%>%
  group_by(country) %>%
  summarise(rai = mean(n_RAI, na.rm = TRUE))

global_decentralization<-global_decentralization %>%
  left_join(rai, by="country")

## Collapse political data

qog <-read_csv("data/qog_std_ts_jan20.csv") %>%
  filter(year>1979 & year<2010) %>%
  rename(country=cname) %>%
  select(ccode, country)

library(texreg)
m1<-lm(rai~oul, global_decentralization)
m2<-lm(rai~share_roufe, global_decentralization)
m3<-lm(rai~share_roufm, global_decentralization)
screenreg(list(m1,m2,m3), 
          stars = c(0.01, 0.05, 0.1))

ggplot()

m1<-lm(political~oul, global_decentralization)
m2<-lm(political~share_roufe, global_decentralization)
m3<-lm(political~share_roufm, global_decentralization)
screenreg(list(m1,m2,m3), 
          stars = c(0.01, 0.05, 0.1))

m1<-lm(fiscal~oul, global_decentralization)
m2<-lm(fiscal~share_roufe, global_decentralization)
m3<-lm(fiscal~share_roufm, global_decentralization)
screenreg(list(m1,m2,m3), 
          stars = c(0.01, 0.05, 0.1))

m1<-lm(administrative~oul, global_decentralization)
m2<-lm(administrative~share_roufe, global_decentralization)
m3<-lm(administrative~share_roufm, global_decentralization)
screenreg(list(m1,m2,m3), 
          stars = c(0.01, 0.05, 0.1))

m1<-lm(dec_index~oul, global_decentralization)
m2<-lm(dec_index~share_roufe, global_decentralization)
m3<-lm(dec_index~share_roufm, global_decentralization)
screenreg(list(m1,m2,m3), 
          stars = c(0.01, 0.05, 0.1))

fig6a<-global_decentralization %>%
  ggplot(mapping=aes(x=share_roufe, y=rai)) +
  geom_smooth(method = "lm", color= "red") +
  geom_smooth() +
  theme_bw() +
  labs(title="Vote-Buying",
       y="Decentralization (RAI)",
       x="Share of Parties Relying on Unions")

fig6b<-global_decentralization %>%
  ggplot(mapping=aes(x=share_roufm, y=rai))+
  geom_smooth()  +
  geom_smooth(method = "lm", color= "red")+
  theme_bw()+
  labs(title="Vote-Monitoring",
       y="Decentralization (RAI)",
       x="Share of Parties Relying on Unions")
```


```{r,eval=FALSE,  echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, fig.cap = "Social Policy Conditions as Share of Total IMF Loan Conditions", fig.width=5, fig.height=3, cache= F}
fig6<-grid.arrange(fig6a,fig6b, ncol=2)
```

```{r, regressions,include=FALSE}
## Bayesian linear regression


bayes.hm <- stan_glmer(education_decentralization_reform~completion_rate +
                        right_center_of_gravity + imf_education_loans+oul  + (1|countryname),
                    data = pol,  gaussian(link = "identity"),
                    chains = 4, seed = 123, iter= 10000)

prior_summary(bayes.hm)

## Get the posterior in good shape

posterior1 <- as.matrix(bayes.hm)
posterior1<- posterior1[,c(1:6)]
colnames(posterior1) <- c("Intercept", "Completion Rates", "Right CoG",
                          "IMF Loans", "OUL", "Polity")

## Draw the Figure

fig4<-mcmc_intervals(posterior1,
           pars = c("Completion Rates", "Right CoG", "IMF Loans", "OUL",
                    "Polity")) +
            theme_bw()+
            labs(y="",
                 x="",
                 caption = "Medians and 95% Intervals",
                 title = "Posterior Distributions")
fig4
```

```{r, plotfig4, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, fig.cap = "Association Between Education Decentralization and Opposition-Union Link", fig.width=5, fig.height=3.5}
fig4
```

```{r, diagnostics, eval=FALSE, include=FALSE}
## Convergence Diagnostics for the Appendix

## Load packages
library(ggmcmc)
library(socviz)
library(hrbrthemes)
library(myriad)

## Get the posterior 
post1<-ggs(bayes.hm)
levels(post1$Parameter)

P <- data.frame(
  Parameter=c("beta[1]","beta[2]","beta[3]","beta[4]","beta[5]"),
  Label=c("Completion Rates", "Right CoG", "IMF Loans", "OUL", "Polity"))

part1<-ggs_density(ggs(bayes.hm, par_labels=P, family="beta")) + labs(subtitle="Posterior Distribution") +
  theme_ipsum_rc()+
            theme(panel.border = element_rect(fill = NA, size = 1),
                  text = element_text(size = 12),
                  axis.title.x = element_text(size = 12, hjust = 0.5),
                  axis.title.y = element_text(size = 12, hjust = 0.5)) 
ggsave("figures/fig_density.png", plot=part1,  height=10, width=8, units="in")


part2<-ggs_traceplot(ggs(bayes.hm, par_labels=P, family="beta")) + labs(subtitle="Traceplots") +
  theme_ipsum_rc()+
            theme(panel.border = element_rect(fill = NA, size = 1),
                  text = element_text(size = 12),
                  axis.title.x = element_text(size = 12, hjust = 0.5),
                  axis.title.y = element_text(size = 12, hjust = 0.5)) 

ggsave("figures/fig_trace.png", plot=part2,  height=10, width=8, units="in")

part3<-ggs_geweke(ggs(bayes.hm, par_labels=P, family="beta")) + labs(subtitle="Geweke Diagnostics") +
  theme_ipsum_rc()+
            theme(panel.border = element_rect(fill = NA, size = 1),
                  text = element_text(size = 12),
                  axis.title.x = element_text(size = 12, hjust = 0.5),
                  axis.title.y = element_text(size = 12, hjust = 0.5)) 
ggsave("figures/fig_gew.png", plot=part3,  height=5, width=8, units="in")
```




Figure 4 summarizes the posterior for the principal explanatory variables. Full model results and evidence regarding convergence are available in section B.5, page 11 of the appendix. The estimates indicate that after seeing the data that the presence of an OUL is associated with an increase of about 11 percentage points in the probability of a decentralization reform in a given administration, the Bayesian confidence interval suggests that there is about a 95% probability that the effect is positive.

Most other covariates are not associated with reform. Examining them in turn, small increases in completion rates yield very small reductions in the probability of education reform. While the direction of this estimate is in line with conventional wisdom—better systems are less likely to reform—the effects are not substantively meaningful: we would expect that an increase of 10 percentage points in completion rates would reduce the probability of reform by less than 5 percentage points. The marginal posterior for the partisan control of government is centered around 0, with a wide standard deviation, suggesting that right-wing governments are no more likely than those on the left to pursue education decentralization reform. This is in line with the expectation that “right” and “left” are weak labels in Latin America.

The number of education sector loans is positively associated with education reform, and the effect is large. The posterior is centered around 22 percentage points, with essentially zero probability of the estimate being negative. While the magnitude is large, I interpret this coefficient with caution given concerns over simultaneity. Specifically, governments that decide to reform education are more likely to request loans and technical assistance from either the IMF or the World Bank. In short, while it is possible that loans shape reform, it is equally likely that reform shapes loans. Models without this term produce similar posterior estimates. Finally, as expected, the level of democracy is unrelated to education decentralization reform.

```{r, figure5, include=FALSE}
## Predicted Probabilities

## Load pakacge
library(BayesPostEst)

## Run Bayesian logit regression 
bayes.hm.logit <- stan_glmer(education_decentralization_reform~completion_rate +
                        right_center_of_gravity + imf_education_loans+oul +
                        polity + (1 | countryname),
                    data = pol,  family = binomial(link = "logit"),
                    chains = 4, seed = 123, iter= 10000)

## Get the results 
mcmcReg(list(bayes.hm,bayes.hm.logit),
        format = 'latex', doctype = F)

## Prepare posterior
mcmcmat.rstanarm <- as.matrix(bayes.hm.logit)

## Set the logit model for prediction
mm <- model.matrix(education_decentralization_reform~completion_rate +
                        right_center_of_gravity + imf_education_loans+teachers_with_opp +
                        polity + as.factor (countryname),
                   data = pol)

aveprob.teachers.jags <- mcmcAveProb(modelmatrix = mm,
            mcmcout = mcmcmat.rstanarm[, 1:ncol(mm)],
            xcol = 5,
            xrange = c(0, 1),
            link = "logit",
            ci = c(0.025, 0.975),
            fullsims = TRUE)

library("ggridges")

fig5<-ggplot(data = aveprob.teachers.jags, 
       aes(y = factor(x), x = pp)) + 
  stat_density_ridges(quantile_lines = TRUE, 
                quantiles = c(0.05, 0.5, 0.95), vline_color = "white") + 
  scale_y_discrete(labels = c("No Teachers with Opposition", "Teachers with Opposition")) + 
  ylab("") + 
  xlab("Estimated Probability of Education Reform") + 
  labs(title = "Probability Based on Average-Case Approach")+
            theme_bw()
```


```{r, plotfig5, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, fig.cap = "Posterior Predicted Probability", fig.width=8, fig.height=4}
fig5
```

Figure 5 shows how the presence of an opposition-union link relates to the predicted probability that an average presidential administration will decentralize education. To do so, I calculate the predicted probability ($Pr(y=1)$)  at pre-defined values of OUL, either 0 or 1, while all other covariates are held at a “typical” value, following suggestions by @king_making_2000. The average probability of reform for any given administration is relatively low, which makes sense given that while a majority of Latin American countries engaged in reform, only a handful of administrations advanced it. Results show a relatively large increase in the predicted probability of reform—about 10 percentage points. The figure implies that there is a large probability that after seeing the data, countries that have an OUL have a higher probability of engaging in education reform than those that do not.


The estimates in this section support the notion that electoral politics influence decentralization efforts. That electoral variables matter above and beyond technical considerations, ideological orientation, and democracy says something powerful about the limits of democratic rule. While others have shown that electoral competition can lead to an expansion of the provision of public services as parties seek to capitalize on the votes of the underprivileged, my analysis implies that electoral competition can lead incumbent political parties to make substantive governance changes to social services that weaken the organized civil society groups. Of course, since the data are observational, these are tentative conclusions. However, together with the examination of El Salvador and Paraguay, the analysis is strongly suggestive of the fact that reforms to education can proceed on the backs of teacher mobilization.

# Conclusion

That politics structures education systems is an increasingly important point in the small literature of education [@ansell_political_2013; @kosack_education_2012; @moe_comparative_2016]. My work moves this literature forward by making explicit that changes to education systems can serve as an electoral strategy to weaken political opponents. My work also asks us to reassess conventional wisdom regarding the link between electoral competition and the provision of public services. Where other authors have highlighted how electoral competition can lead to an expansion of social policy  [@dion_international_2008;@garay_social_2016;@de_la_o_crafting_2015], I show that electoral competition can lead incumbent political parties to make targeted changes to social policy to hurt the preexisting organizational and mobilizational advantages of their rivals. 

My analysis offers important public policy lessons. A robust literature on education reform emphasizes the role that teachers’ unions play in blocking reforms [@bonilla_pensiones_1998; @corrales_politics_1999; @graham_improving_1999; @moe_comparative_2016; @diaz_rios_when_2016]. Yet the argument developed here asks us to reassess the commitments of elected leaders to improve education above all else. The simple point that incumbents want to win elections has implications for how we understand why education reforms are successful. We should expect, for instance, that decentralization efforts are least likely to be pursued in good faith in the presence of an opposition-union link, because the overriding concern of parties should be to weaken the opposition. Such reforms, however, are positioned to fail, both because they will never get the buy-in of teachers and because a return to power by the opposition will dismantle them (such as in the case of El Salvador). Conversely, changes to the education system should be most likely to stick when teachers are unattached to political parties because incumbents are motivated to reform for reasons other than weakening opponents, leading them to engage unions extensively to do so.  

\theendnotes
