---
title: 'Decentralization as a Political Weapon: Education Politics in El Salvador and Paraguay^[Forthcoming in *Comparative Politics*.]'
author: "Thomas R Vargas^[Postdoctoral Fellow, Department of Political Science, McMaster University, Hamilton, ON. Corresponding address: vargasct@mcmaster.ca. I thank Teri Caraway, Michelle Dion, John Freeman, Claire Le Barbenchon, Dave Lopez, Oanh Nguyen, Oscar Pocasange, David Samuels, Ben Ross Schneider, and two supportive and thorough anonymous reviewers for their comments on previous drafts of this paper. I am also grateful to seminar participants at the University of Minnesota and Duke University, and audiences at APSA 2019 and MPSA 2018. My thanks to the University of Minnesota for research support and to the World Bank Archives for access to key material. Any remaining mistakes are my own.]"
abstract: \singlespacing What explains why some governments advanced decentralized education in the 1990s while others shied away from such efforts? Some arguments suggest that decentralization was pursued to improve the coverage and quality of education. Others point to partisanship, ideology, or diffusion. Drawing on a case study of El Salvador and Paraguay, I argue instead that governments pursued education decentralization in part because it could be deployed as a political weapon to weaken teachers’ unions affiliated with the opposition, thus depressing mobilization and votes for their rivals. These findings contribute to the literature on decentralization by highlighting a new political motivation fueling decentralization efforts across the developing world—the demobilization of the opposition.
date: "This draft: September 22, 2021"
bibliography: setup/bib.bib
mainfont: Times New Roman
sansfont: Times New Roman
colorlinks: true
documentclass: article
geometry: margin=1in
fontsize: 12pt
linestretch: 1.5
link-citations: true
output:
  pdf_document: 
        includes:
          in_header: setup/header.tex
        number_sections: true
  html_document: default
  word_document: 
    reference_docx: setup/word-template.docx
---
\newpage

# Introduction

Public investments in primary education have huge payoffs in terms of long-term development outcomes. Many countries in the developing world, however, struggle to improve the quality and coverage of public education. One widely touted reform designed to improve such outcomes is *decentralization*^[I define decentralization as the transfer of administrative, fiscal, or political power from higher levels of government to lower levels of government. Education decentralization as understood in this manuscript is a special case of administrative decentralization, where choices about the management of schools are devolved to lower levels of government (such as municipalities or local school boards).]. In this regard, scholars are presented with a puzzle: some countries around the world with low educational achievement and coverage rates advanced the decentralization of primary and secondary education, while others with similar attributes did not. If—as prominent observers have suggested[@grindle_despite_2004; @kaufman_politics_2004]—governments pursue decentralization in order to improve education (a “technical” goal), the explanation for such variation remains a mystery. I argue, by contrast, that governments often wield decentralization as a weapon against the opposition whenever it is allied with teachers’ unions.

Consider the cases of El Salvador and Paraguay. In 1991 the Salvadoran government pushed for a school-based management program called *Educación con Participación de la Comunidad* (EDUCO) that would eventually devolve the governance and administration of most rural schools to local communities. Under EDUCO, rural communities would oversee infrastructure, collect school fees, and hire teachers—a truly extensive and radical rethinking of how to manage the provision of public education in the region. By contrast, Paraguayan authorities did not push for the decentralization of education despite their country’s many similarities to El Salvador during its EDUCO years: it had low primary education completion rates, it was beset by high levels of poverty in rural areas, and it was ruled by a right-wing government (for other similarities, see table 1). While Paraguayan elites were aware of decentralization programs elsewhere, they chose not to pursue them.

The difference in support for decentralization projects can be explained by the partisan affiliations of teachers. Following the Salvadoran civil war (1980–1992), the incumbents faced considerable uncertainty regarding their electoral prospects under democracy. The ruling right-wing party, *Alianza Republicana Nacionalista* (ARENA), feared losing power to a guerrilla-turned-legitimate political actor, the *Frente Farabundo Martí para la Liberación Nacional* (FMLN). To avert the electoral success of the left, ARENA sought to limit the ability of teachers’ unions to mobilize in favor of the FMLN, their political allies, by pursuing education decentralization. In Paraguay the absence of a substantive link between teachers’ unions and the opposition party meant there were no political incentives to reform the governance of education, even though the education system’s performance (measured by completion rates, for instance) was as weak as El Salvador’s.

These examples demonstrate the political logic that can help explain why some Latin American governments pursued reforms to education governance in the 1990s while others did not: incumbent political parties supported decentralizing reforms because their second-order effects made it more difficult for teachers’ unions affiliated with opposition parties to mobilize in their favor. This argument is developed by bringing together the extant literature on the electoral advantages of political and fiscal decentralization for incumbents [@montero_decentralization_2004;@oneill_decentralization_2003;@oneill_decentralizing_2005; @niedzwiecki_social_2016]  with scholarship exploring the partisan identities of teachers’ unions [@larreguy_political_2017;@moe_comparative_2016; @murillo_recovering_1999]. 

This study contributes to our understanding of decentralization politics in two ways. First, I document a new political motivation for decentralization. Previous work has shown that incumbent parties undertake decentralization when their support at the national level is waning [@oneill_decentralizing_2005; @dickovick_municipalization_2007; @gonzalez_political_2008].  The logic is that national incumbents are more competitive in the newly empowered subnational units. In the argument developed here, political incumbents *reshape* the existing (national) arena to their advantage by targeting opponents and their allies. This theory draws on work by Hertel-Fernandez -@hertel-fernandez_policy_2018 in the American states showing that conservative groups advanced policies whose knock-on effects demobilized political opponents. The same underlying political logic can help explain why seemingly nonpartisan education decentralization efforts were pursued with such fervor in certain contexts and ignored in others.

Second, this paper challenges the idea that decentralization can be thought of as a wholesale process entailing the concurrent devolution of administrative, fiscal, and political responsibilities.  In Falleti’s -@falleti_decentralization_2010 classical framework, the decentralization of education (which she calls an “administrative” reform) cannot be separated from devolution projects in other domains (fiscal and political). In this piece I establish that administrative decentralization with no corresponding reforms shaping political or fiscal authority is possible, that it is not toothless, and that it can accomplish significant political goals. This finding builds on previous work showing that political decentralization can be advanced in the absence of concurrent fiscal or administrative decentralization to further political aims [@bohlken_democratization_2016].  But unlike political decentralization, administrative changes are often couched in “technical” arguments, possibly making them easier for governments to push for. As I show in this piece, this does not preclude them from being partisan. 

Finally, this study pushes forward the literature on education politics. Most of this scholarship emphasizes how the process of education decentralization unfolds, assuming away the reasons incumbents push for decentralization and focusing instead on the strategies teachers’ unions use to limit reform [@grindle_despite_2004;@kaufman_politics_2004;@corrales_politics_1999; @murillo_recovering_1999; @falleti_decentralization_2010].  My dependent variable is not the ultimate success or failure of decentralization efforts, but whether or not incumbents select decentralization as a policy in the first place. In doing so I shift focus slightly away from teachers’ unions. Still, I do not dispute the dominant paradigm: teachers can be powerful players in the politics of education, and their partisan attachments inform the strategies they follow in defending their vested interests [@moe_comparative_2016; @bruns_politics_2019].  But an emphasis on what teachers *do* (specifically, for instance, block reform) understates a central fact that structures democratic politics everywhere: parties want to win elections. Therefore, we should not assume that governments propose politically costly changes to education systems exclusively because they want to improve them. Viewing education reform through this electoral lens can turn received wisdom on its head: the partisan attachments of teachers’ unions, broadly believed to empower them to resist change [@murillo_recovering_1999],  can actually set them up to lose.

# Existing Accounts of Decentralization Motivations

Earlier work on social policy sector reform in Latin America suggests that the origins of education decentralization spring from one of three causes: technical factors, ideational pressures, and diffusion. Each is insufficient to address the puzzle at the beginning of this paper.

## The Technical Explanation

The technical determinants theory holds that some governments in the region decentralized because their public education systems were overwhelmed by a crisis of coverage and quality [@grindle_despite_2004;@kaufman_crucial_2004].  Through the twentieth century, reformers argued, Latin American governments created bureaucratic behemoths that were too slow to meet these challenges. According to this line of thinking, the only viable solution to improving the coverage and quality of education was to break up concentrated authority. This would achieve gains in efficiency at the same time that it democratized decision-making by devolving decisions over education to local communities. In countries with lagging enrollments in faraway rural areas, like Paraguay and El Salvador, decentralization would make it easier to expand schooling without having to go through the national bureaucracy.


```{r, figure1, include=FALSE}
## Load packages
library(tidyverse)

## Load data
qog<-read_csv("data/qog_std_ts_jan19.csv") %>% # QoG dataset
  select("cname", "year", "ccodecow", "ccodealp", # select only relevant variables
                           "wdi_expedu", "bmr_dem", "p_polity2", "wdi_pop14",
                           "gle_cgdpc","gle_trade", "wdi_expeduge", "wdi_expedup",
                           "wdi_expedus", "ht_region",
                           "wdi_expedut", "wdi_expstup", "wdi_expstus", "wdi_expstut",
                           "bl_lpmf" , "wdi_popurb", "wdi_debt", "bl_asymf", "wdi_gerp",
                           "wdi_nerp") %>%
  rename("cowcode" = "ccodecow" ) 

## Merge data 

data<-read.csv("data/TRV_decentralization.csv") %>% # Decentralization data
  left_join(qog, by=c("cowcode", "year")) # Merge with QoG

## Plot performance data

data%>%
  select(full_name, year, wdi_gerp) %>%
  arrange(full_name)%>%
  subset(year>1977)

f1a<-data %>%
  ggplot(mapping=aes(x=year, y=wdi_gerp)) +
        geom_smooth(color="black", size=2) +
        geom_jitter(alpha = 0.2) +
        theme_bw() +
        xlab("Year") +
        ylab("Enrollment Rates") +
        coord_cartesian(xlim = c(1970,2010), ylim = c(0,120), expand = TRUE,
                        default = FALSE, clip = "on") +
        scale_x_continuous(breaks = seq(1970, 2010, by = 10)) 

f1b<-data %>%
  ggplot(mapping=aes(x=year, y=wdi_nerp)) +
        geom_smooth(color="black", size=2) +
        geom_jitter(alpha = 0.2) +
        theme_bw() +
        xlab("Year") +
        ylab("Completion Rates") +
        coord_cartesian(xlim = c(1970,2010),ylim = c(0,120), expand = TRUE,
                        default = FALSE, clip = "on") +
        scale_x_continuous(breaks = seq(1970, 2010, by = 10))

library(gridExtra)

fig1<-arrangeGrob(f1a,f1b, ncol=2)

ggsave(fig1, file="figures/fig1.pdf", height = 3, width = 5)
```

```{r, plotfig1, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, fig.cap = "Enrollment and Completion Rates for School-Aged Children Latin America Since 1970", fig.width=5, fig.height=3, cache= F, label=fig1}
grid.arrange(f1a,f1b, ncol=2)
```

The idea that governments pursued decentralizing reforms exclusively because of these technical concerns does not find much support in the data. As shown in figure 1, during the “lost decade” of the 1980s the region as a whole experienced no reversals in primary school coverage. Indeed, countries with low enrollment rates (like El Salvador and Paraguay) experienced rapid gains. Decentralization was not needed to improve enrollment. Perhaps governments were interested in improving quality issues via decentralization; education quality was generally low across the region. Yet empirical evidence regarding the quality of education was nonexistent until well into the 2000s. As documented by UNESCO in 2001: “we do not yet have a time series of education achievement indicators that is sufficiently broad to adequately describe the behavior and trends of such indicators within countries in the region” [@unesco_state_2001, p.10].  That Latin American governments supposedly concerned with learning didn’t collect more evidence regarding their quality-improving reforms suggests that this was, at best, a secondary goal to decentralization initiatives.

Still, it is plausible that regional trends in decentralization were driven by countries that, on average, trailed their peers. For instance, countries with the lowest enrollment rates (which in the absence of learning data might reasonably stand in for quality issues) may have been under the most pressure to decentralize. This would be consistent with the notion that technical considerations propelled reform. As the cases of El Salvador and Paraguay illustrate, however, there are discrepancies in decentralization decisions even across countries with “weak” education systems (in terms of either enrollment or quality). In short, to account for regional variation in the decentralization of public education, we have to look beyond technical concerns regarding the coverage and quality of schools.

## Ideational Theories 

The ideational theory of education reform attributes decentralization to neoliberal policy ideas that gained traction among center-right and right-wing parties. These parties, coming to power in the 1990s, were keen on introducing market-oriented strategies to the provision of social welfare. In the face of pervasive deficits, these governments were interested in keeping social expenditures sustainable and social policy programs cost-effective [@nelson_social_2011].

For states burdened with debt, passing responsibility for the management and funding of education to subnational units offered an opportunity to reduce central government spending [@falleti_decentralization_2010].  In theory, aggregate education expenditures would also fall, because the cost associated with collecting knowledge about prices and production processes is lowest at the local level [@shah_balance_1998].  Similarly, decentralization was thought to yield innovation in the provision of services [@oates_fiscal_1972; @ford_rationale_2000; @tiebout_pure_1956].  For neoliberal advocates, then, decentralization offered the potential efficiencies and innovation associated with free markets. Yet decentralization has not been consistently associated with a single political-economic logic [@montero_decentralization_2004].  During the time under analysis, both right- and left-wing governments proposed education decentralization programs. Perhaps most interestingly, various right-wing governments have shied away from decentralization. For example, subsequent *Colorado* administrations in Paraguay never seriously considered decentralization.

It’s reasonable to imagine, nonetheless, that it was those states most burdened by the debt crisis that were compelled to decentralize education as a cost-saving mechanism. This is another variation of the technical incentives argument that government officials used in trying to sell decentralization. But countries saddled with high levels of debt in the aftermath of the debt crisis followed different trajectories with regard to education reform. Take again the examples reviewed at the opening of this paper. While both El Salvador and Paraguay experienced sharp increases in debt service as a share of GDP during the 1980s, by 1989 Paraguay’s outstanding external debt accounted for .53 of GDP compared to .50 for El Salvador [@huber_social_2012].  Yet it was El Salvador and not Paraguay that engaged in decentralization efforts. In short, fiscal considerations alone cannot explain education reform.

## Diffusion Explanations

Finally, the diffusion arguments suggest that governments began making certain types of changes to their education systems to accommodate both international financial institutions (IFIs) and changing global norms surrounding schooling. The early 1990s represented a unique time in international education. Donors and international organizations heavily promoted the idea that education was crucial to development. This is best exemplified by the World Conference on Education for All held in Thailand in 1990—financed by international organizations—which set five target education goals for 155 countries. One possible mechanism is that these organizations, including the World Bank and the International Monetary Fund (IMF), pressured governments into adopting their own ideas about “what works” in education by funding certain education reform projects, or by attaching conditionalities to structural adjustment programs (SAPs). We would therefore expect that countries with active SAPs or sectoral loans with IFIs would be more likely than others to decentralize education.

To assess whether this was plausible, I examined 193 IMF loan agreement documents, including letters of intent, arrangement letters, and staff reports, for eighteen Latin American countries from 1980 to 2000.^[Thanks to Teri Caraway for sharing her collection of documents with me.]  I looked for conditions that required education sector reform (of any type) prior to the disbursement of loans— what are known as “hard” conditions [@woods_globalizers_2014].  I found no trace of “hard” conditions placed on the education sector, let alone decentralization projects. I did uncover two instances of what are known as “soft” conditions, conditions used mostly to track progress in program implementation [@caraway_international_2012;@kentikelenis_imf_2016],  in Argentina and Bolivia. These soft conditions, however, were meant to encourage the implementation of decentralization programs already approved by the respective legislature, not the advancement of new decentralization projects. In sum, there is limited evidence suggesting that IFIs coerced or enticed governments into pursuing these types of policies.

On the other hand, international organizations surely helped shape the conversation around education reform in the 1990s [@edwards_jr_trajectory_2017;@mundy_world_2016].  IFIs may have relied on their convening power to instill in politicians and high-level bureaucrats the importance of decentralization. While this is plausible, conventional wisdom tends to overstate how emphatically agencies like the World Bank advocated for education decentralization specifically. In working papers widely circulated around the Bank in the 1980s and the mid-1990s, Bank staff warned that the decentralization of public services was no panacea, especially in poor countries with weak states [@rondinelli_decentralization_1983; @prudhomme_dangers_1995].  My exploration of recently declassified internal World Bank correspondence related to various education reform efforts in El Salvador and Paraguay suggests that the Bank was more interested in increasing expenditures in primary education than in advocating for any particular education governance reform. It remains possible that individual IFI staff advocated heavily for decentralization, convincing mid-level bureaucrats and some politicians that governance reforms were necessary to fix some quality or coverage issue. But this does not explain *why* governments finally decided to expend political capital on pursuing a costly endeavor. Thus, while ideas about decentralization were surely available and known to bureaucrats and politicians, their availability cannot have been the principal *driver* of reform.

# Decentralization as a Political Weapon to Demobilize the Opposition

My argument is that the partisan affiliation of teachers determines whether or not incumbents push for education decentralization. In competitive electoral environments, incumbents win power through elections. In such settings parties often rely on linkages with large civil society organizations—like teachers’ unions—to drum up votes. When incumbents identify a linkage between teachers and the opposition, they will search for ways to weaken teachers’ organizational advantages to hobble opponents. Education decentralization offers incumbents a powerful tool to do so. It is well known that policies shape politics in unintended ways through policy feedback processes [@soss_lessons_1999;@skocpol_protecting_1995;@pierson_when_1993;@campbell_how_2003].  I build on the notion that politicians deliberately (try to) “make politics” by advancing policies that will improve their electoral position down the road [@anzia_politicians_2016;@hertel-fernandez_policy_2018].  This is an idea already present in the decentralization literature. O’Neill -@oneill_decentralizing_2005,  for instance, has argued that incumbent political parties give up centralized fiscal authority to subnational units whenever two conditions are met: they fear losing power at the national level, and they believe their electoral fortunes are more secure at lower levels of government. More recently Bohlken -@bohlken_democratization_2016 has shown that higher-level government elites use local political decentralization reforms as a means of exerting control over lower-level intermediaries. In both cases incumbents choose political and fiscal decentralization because, counterintuitively, it strengthens their electoral or political position.

I consider instead how governments use administrative decentralization to kneecap their *rivals*, a strategy @hertel-fernandez_policy_2018  calls *policy as political weapon*. Incumbents decentralize education not because they want to withdraw from national elections or better monitor their clients, but because doing so weakens their electoral opponents through knock-on effects. In Hertel-Fernandez’s policy-as-weapon framework, politicians have three paths to demobilization. First, they might construct policy to make it more difficult for opposing interest groups to attract and hold members. Second, they might generate barriers to civic engagement to the detriment of the opposition, such as through voting restrictions. Third, they might hobble state capacity so as to limit the ability of political opponents to reap electoral rewards from certain policies. As laid out below, education decentralization as a political weapon targets teachers’ mobilization advantages to limit how they can identify and mobilize voters to support their partisan allies.

I build this argument from a series of insights about the behavior of incumbents, opposition parties, and teachers. To begin we must recognize that democratization changed incentives for political elites across Latin America. Mainly, national elections became the most important way for parties to access political power [@mainwaring_juan_1997].  Because in most countries democratization generated true electoral competition, substantial uncertainty was introduced into the political process. Incumbents did not know whether they would win the next election. Losing would diminish their ability to shape policy and reap rewards for themselves. Therefore, to the degree that policies determine electoral outcomes, successful incumbents will select policies so as to ensure their survival [@downs_economic_1957; @riker_theory_1984]. 

Scholars of social policy presume that incumbents focus on high-profile policies to reward core constituencies or broaden their electorate [@garay_social_2016; @fairfield_redistribution_2017].  But in some, perhaps most, democracies (and certainly in rich ones as well) incumbents leverage whatever policy tools they have at their disposal to shift the playing field their way. Aside from courting voters by targeting policies to supporters, incumbents are likely to advance policies that weaken their electoral opponents, including the implementation of voting restrictions and result manipulation [@mares_buying_2016; @hafner-burton_when_2014].  In sum, incumbents and parties are strongly motivated by elections [@mayhew_congress_1974;@samuels_ambition_2003],  and we should view costly, highly visible changes to social policy as a potential opportunity to hobble their opponents.

A second point is that in Latin America parties have historically relied on unions to both collect information on voters and deliver votes [@stokes_brokers_2013; @szwarcberg_uncertainty_2012; @lowden_education_2004; @eaton_teachers_2014].  Teachers are uniquely positioned to do so because of their distribution, numbers, and position of authority. Indeed, across much of the developing world, there is mounting evidence that teachers act as political brokers and that their unions are caught up in complex webs of clientelism and patronage [@sandholtz_voters_2019;@beteille_absenteeism_2009;@chattharakul_thai_2010;@pierskalla_personnel_2019;@eaton_teachers_2014].   In Colombia, for instance, up until recently it was typical for teachers to be hired en masse around election time [@lowden_education_2004]. In Mexico many teachers affiliated with the Mexican National Educational Workers’ Union (SNTE) similarly owed their jobs to Partido Revolucionario Institucional (PRI) leaders, who weaved a web of patronage-based practices regarding the hiring and management of teachers [@chambers-ju_teachers_2017].  Some have argued that the PRI’s electoral victories were secured in part by the mobilizing efforts of the SNTE and that these attachments persist [@larreguy_political_2017].  If teachers form an important part of the opposition’s electoral efforts, then incumbents are likely to take steps to weaken them.

My argument doesn’t require that teachers’ unions be fully co-opted by parties and unable to occasionally act independently. Indeed, teachers’ unions sometimes offer criticism of their partisan allies and threaten strikes against them. For example, in Brazil in 2000 and 2004 the *Centro dos Professores do estado do Rio Grande do Sul* went on strike against the Workers’ Party (PT) despite their close electoral alliance [@goldfrank_restraining_2003]. At the same time, these linkages do not need to be permanent. In the aftermath of neoliberal reform and democratization, the ties between teachers’ unions and parties underwent considerable transformation [@chambers-ju_adjustment_2021].  My assumption, however, is that a perceived close connection between the opposition and teachers is sufficient to provide an incentive to use policy as a weapon to weaken teachers.

A third insight is that in centralized systems the costs of education decentralization fall disproportionately on teachers.^[In cases where teachers do not bargain with the national government, like in Brazil, decentralization does not represent as much of a threat and must be explained by other logics.]  Teachers have organizational and mobilizational advantages over most public employees. Not only do teachers typically outnumber every other public employee group by a fair margin [@bruns_great_2015],  but they are also typically dispersed across all areas of a country and are deeply integrated into local communities [@moe_comparative_2016].  These advantages are compounded by national collective bargaining because it makes organizing into one or a few unions easier. When the teacher movement is cohesive, teachers can credibly threaten to paralyze a country when a government fails to meet their demands. When education systems shift to decentralized governance, however, these advantages weaken because a single teachers’ union now must contend with multiple employers. More specifically, as a result of decentralization teachers lose cohesion—because teachers face differing hiring and incentive structures across a national territory—and therefore negotiating leverage, at least with the central government. The natural outcome is a fracturing of teachers’ unions. In some cases, like the one below, it may mean the exclusion of unions from certain subnational territories. In turn, teachers expect to face lower wages, less access to jobs, and worsening working conditions [@hanson_educational_1997; @murillo_recovering_1999].  For these reasons, teachers’ unions tend to strongly oppose decentralization [@grindle_despite_2004; @kaufman_crucial_2004], sometimes to the point of violence [@bruns_managing_2016].  Thus, decentralization can be used as a weapon against teachers, and by extension, their partisan allies.

To summarize, incumbents will advance a politically costly decentralization project only if they have strong electoral incentives to do so. The logic is as follows. The overarching interest of incumbents is to maintain power. In democracy incumbents remain in power through regular elections. Because teachers are a highly organized mobilized group, with deep and sometimes opportunistic partisan attachments, they form the backbone of many electoral coalitions. However, the power of unions depends on cohesion and reach. Decentralization, through the weakening of national collective bargaining and by limiting the size of unions, affects the mobilizational advantages of unions. If unions are weak, then they are less likely to be able to deliver votes for their electoral coalitions. Incumbents know this, in part because they draw on similar ties to mobilize votes. As a result, incumbents will pursue education decentralization as a political strategy whenever teachers are allied with the opposition. Whether or not decentralization is implemented and whether it actually changes the distribution of power across levels of government depends on processes already described in the literature [@falleti_decentralization_2010;@niedzwiecki_social_2016; @montero_decentralization_2004; @grindle_audacious_2000], but cannot be fully understood without examining why incumbents invest in these programs in the first place.

A final caveat is in order. My argument does not rely on decentralization *actually* weakening the opposition in the ways hypothesized here. The policy feedback process is complex and always generates intended and unintended effects. For example, it is conceivable that decentralization projects might have united opposition groups in powerful and enduring ways they themselves could not have orchestrated, leading to subsequent successes at the ballot box. Or perhaps teacher mobilization was never going to be decisive in an upcoming election. My point is that politicians have ideas about how certain policies might advantage them down the road, and identifying whether or not those incentives exist can go a long way toward revealing the political logic of decentralization.

# Case Study Evidence

To draw the logic of this theory I look at the case of El Salvador, described briefly at the opening of this paper, with reference to the contrasting case of Paraguay. Case selection follows a most-similar-systems design: despite their structural and political similarities, these countries differ on outcomes (see table 1). El Salvador pursued a broad-scale decentralization program that deeply transformed patterns of authority across public education, whereas Paraguay engaged in various education sector reforms that never truly decentralized education. By holding constant structural characteristics commonly used to explain why countries engaged in decentralization, I show that these are not sufficient to explain patterns of education governance reform. The argument developed above can explain this puzzle.

One notable difference is the fact that El Salvador was emerging from a civil war, while Paraguay was not. I argue that this is not a problem for the paired comparison because in both countries social and political actors wished to turn the page on a violent past in rural areas. In El Salvador my interviews with political leaders confirmed that both incumbents and revolutionary leaders were eager to end fighting and were optimistic about their chances to influence politics through elections ^[Interview ES7, ES14].  In Paraguay the Stroessner and Rodríguez regimes were notable for conflicts in the abandoned rural areas. The transition process, however, also generated optimism among political actors that violence was behind them [@nagel_unleashing_1999].  In neither case was the “return” of abandoned rural territories into the national fold in question. More importantly, for my purposes, previously outlawed or weakened opposition groups were perceived as being on the cusp of obtaining political power. While differences remain, I view the transition process of both countries, occurring in a context of optimism about democracy and under true electoral competition, as analytically equivalent.

\input{tables/table1.tex}

The analysis draws on four weeks of fieldwork in El Salvador, including interviews and reviews of historical material, two trips to the World Bank archives in Washington, DC, to assess documents that contained the history of EDUCO in the context of World Bank loans, and secondary sources. An interview methods section can be consulted in section A of the appendix. The case study covers events that happened nearly thirty years ago, placing straightforward limitations on the data—key actors may no longer be alive, or they may adjust their recollections of events to fit dominant narratives. To overcome these issues, wherever possible I triangulate interview data with archival data and secondary materials. The case of Paraguay is constructed mostly from secondary sources and a review of primary documents.

## El Salvador: Education Decentralization as a Political Weapon

From 1940 to 1992 the provision of public education in El Salvador was strongly centralized, with the Ministry of Public Education (hereafter, the Ministry) controlling the establishment, funding, and administration of public schools [@aguilar_aviles_vistazo_1998;@segovia_flores_historia_2012].  The Salvadoran civil war (1979–1992), however, severely affected the provision of public education in rural areas because government forces targeted teachers, who they assumed to be supportive of the FMLN.^[Interviews ES1, ES3, ES6]   As a former guerrilla member put it: “The guerrilla [forces] destroyed the economic infrastructure while the [government] army destroyed the education infrastructure; that was the division of labor.”^[Interview ES14]  Moreover, reports by the @world_bank_salvador:_1990 and @unesco_servicios_1990 pointed to low completion rates and high repetition rates in public primary education in a context of falling central government investment in education. Thus, the Salvadoran government faced a typical problem for developing countries: improving the quality and coverage of education in a context of falling revenue.

Changing political circumstances opened a window for education reform. In 1989 Salvadorans elected the (right-wing) ARENA leader Alfredo Cristiani with a mandate to end the ongoing civil war. In anticipation of the signing of the Peace Accords (in 1992), government leaders were eager to return a sense of normalcy to daily life. This included returning official education services to rural areas and improving the quality of services.^[Interview ES14]  A critical goal of any education reform was to “gradually decentralize management responsibilities to the regional and local levels” [@world_bank_staff_1993 pp.21]  enshrined in the administration’s Socioeconomic Development Plan for the years 1990–1994. The idea was that decentralization, by virtue of delegating authority to lower levels of government, would create efficiencies in the provision of education while improving the quality of education since parents would have a bigger say in the administration of schools and supervision of teachers. In sum, decentralization efforts were couched in technical considerations.

Government efforts to decentralize education were spearheaded by the Minister of Education at the time, Cecilia Gallardo, a hardline conservative with strongly held neoliberal views [@edwards_jr_trajectory_2017].  Gallardo’s menu of policy options was constrained by her ideology and dominant ideas in international education policy circles, both favoring some sort of decentralization strategy. More pressing, however, were the emerging political realities. During the civil war local communities organized locally run schools known as *escuelas populares* to restore their access to education in places abandoned by the government. These *escuelas populares* served the guerrillas’ ideological and political objectives. Ideologically, schools emphasized a state of war against injustice and inequality. Madre Rosa, a teacher, recounts: “I consider myself a *guerrillera* [guerrilla fighter] because [by teaching] we are fighting against a system that oppresses the poor and has never allowed them access to school, not even to learn their ABCs” [@de_la_cruz_educacion_1983 pp.95].  With the civil war coming to an end, it became imperative to exclude leftist elements and their institutions from areas previously controlled by the guerrillas.


The political salience of education reform was highlighted by the upcoming elections. Under the terms of the peace agreement, the FMLN would be allowed to participate as a political party in free elections in exchange for dismantling its military structure [@de_soto_implementation_1995]. In the run-up to the 1994 elections, there were thus two major political parties: ARENA and the FMLN. ARENA believed the elections were going to be competitive and viewed the FMLN as a real political contender.^[The FMLN did not nominate its own presidential candidate, but it did join a coalition of parties to present a nominee. It also ran its own candidates for the legislature.]  As reported by @lehoucq_election_1995, “[u]ncertainty about the electoral strength of the Left unnerved many in government, the military and their conservative supporters” (181). These groups acted accordingly. ARENA-affiliated death squads implemented a low-level campaign of violence against FMLN leaders. Prior to the elections six leaders of the FMLN political coalition were murdered by right-wing death squads. ARENA also took to the airwaves, warning voters that an FMLN victory would threaten the peace and imply a “return to the past” [@stahler-sholk_salvadors_1994].  In addition, there were widespread claims of irregularities in the election process [@lehoucq_election_1995].

This political context affected the strategies pursued by the government to reform public education. Extending public schooling to rural areas could have been achieved by using the existing education governance system, but education leaders thought it would be too slow.^[Interview ES11]  Most importantly, public school teachers and their organizations were viewed with suspicion. Most public school teachers belonged to the *Asociación Nacional de Educadores Salvadoreños, 21 de junio* (ANDES) teachers’ union. The union was viewed as a tool of the radical left with deep partisan attachment to the opposition. To some, the histories of ANDES and the FMLN are one and the same. In 1970 the Secretary General of ANDES, Mélida Anaya Montes, co-founded one of the five component groups of the FMLN. Thus, in my interviews it was common for individuals to lump both organizations together. As a prominent right-wing politician told me, “The guerrilla is *born* in the teachers’ unions. The first guerrilla fighters of El Salvador are teachers.”^[Interview ES10]   While most ANDES-affiliated teachers were not directly involved in the conflict, there is some evidence that their organization provided support to the FMLN through union dues. In a cable from 1986 the CIA reports that Miguel Castellanos, a high-ranking FMLN defecting officer, stated that for all intents and purposes FMLN controlled ANDES, and that about half of all economic support it received was funneled to the guerrillas.^[``El Salvadors’ Insurgents: Resurrecting an Urban Political Strategy” Copy 181, from September 1986 held in the CIA Archives. Sanitized Copy Approved for Release 2011/12/30 accessed on January 21, 2019 available: [here][link1].]

[link1]: https://www.cia.gov/library/readingroom/docs/CIA-RDP04T00794R000100260001-0.pdf

The organizational closeness between the FMLN and ANDES makes clear that the latter’s mobilizational advantages were key to the opposition’s electoral strategy.@sprenkels_after_2018,  who entered Chalatenango in 1992, recalls: “All the communities participated in party organizing, practically down to the last family.” In places where the guerrillas did not have a visible presence, they exerted control through other agents; for example, “the ANDES teachers’ union had departmental (subnational) organizational networks in Chalatenango within which the [guerrilla] maintained strong contacts” (88). In the run-up to the 1994 elections teachers formed a large part of the FMLN’s party infrastructure. Most importantly, education leaders were painfully aware of this union-opposition bond and sought to disrupt it. In a 1991 interview Cecilia Gallardo, then Minister of Education, complained about the unions’ political activity: “What ANDES should do is put on the FMLN’s t-shirt, to which they have always belonged by the way, and go campaigning,”^[From a newspaper article in 1991. “Una crisis de libros y lectores: Licenciada de Cano," La Prensa Gráfica (February 11, 1994).]   implying that the teachers’ union was merely an extension of the FMLN, and that their organizational advantages were used not only to secure the interests of their affiliates but also to support their partisan allies.

The dilemma facing the government, then, was how to expand schooling to rural areas without leaving behind ANDES-affiliated actors who could use their position to identify and mobilize voters for the FMLN. The solution was an administrative decentralization program that replaced *escuelas populares* in rural areas and targeted teacher organization. ARENA and Gallardo aggressively advanced EDUCO, a program under which schools were managed by elected parent councils, although the central government maintained authority over naming principals, funding schools, and setting the curriculum. The program was announced to much fanfare and with the full support of government in 1992.^[Interview ES11] 

The dilemma facing the government, then, was how to expand schooling to rural areas without leaving behind ANDES affiliated actors that could use their position to identify and mobilize voters for the FMLN. The solution was a administrative decentralization program that replaced *escuelas populares* in rural areas and targeted teacher organization. ARENA and Gallardo aggressively advanced EDUCO, a program under which schools were managed by elected parent councils, although the central government maintained authority over naming principals, funding schools, and setting the curriculum. The program was anounced to much fanfare and with the full support of government in 1992.^[Interview ES11]

Two features of the program design reveal this political logic. First, EDUCO was planned to most heavily focus on rural areas and, specifically, FMLN strongholds. As noted by a politician: “The EDUCO model served as a way to transition the guerrilla territory [from the FMLN to the government].”^[Interview ES11]  In the period between 1991 and 1993 EDUCO would quadruple in size by targeting communities precisely in FMLN-controlled areas. On the eve of the 1994 elections EDUCO covered roughly 2,300 teachers and 74,000 students (about 17% of all primary school children), almost all of them in rural areas  [@ornelas_education_2019].  

Second, EDUCO was conceived to weaken teacher organization in these communities by making it as unappealing as possible for unionized members to apply for jobs at EDUCO schools and by making it likely that they would not get hired in the first place. Under EDUCO, teachers would be hired on yearly contracts and would be paid less than teachers in the official system.^[Eventually the World Bank, seeing EDUCO teachers jump ship whenever they could, advocated for a raise in the EDUCO pay schedule. This was after the 1994 elections. One of the unanticipated ramifications of the creation of parallel teaching tracks was that every EDUCO teacher wanted to enter the non-contracting option (Interviews ES1, ES6, ES12).]  Teaching in EDUCO schools was therefore not appealing to tenured teachers affiliated with ANDES.^[Interviews ES5, ES6, ES12]  At the same time, EDUCO teachers would be required to meet a minimum set of teaching qualifications (i.e., obtaining some sort of degree). This was meant to exclude instructors who had been teachers in escuelas populares.^[To allay the concerns of parents who had warmed up to teachers in escuelas populares, the Ministry of Education offered these teachers a path to formalization (Interview ES12). Instructors who had taught in escuelas populares would be eligible to teach under EDUCO if they underwent special trainings (Interview ES11). In practice, however, this had the effect of limiting the number of teachers in escuelas populares who would become EDUCO teachers, since it entailed leaving their posts, unpaid, for a significant amount of time. While it might be the case that 80 to 90% of teachers accepted the Ministry’s offer, as suggested by a former Ministry official (Interview ES11), the immediate job openings clearly presented a political opportunity for ARENA.]  Finally, available teaching jobs would be doled out to ARENA sympathizers. In theory, the hiring of teachers would run through the parent councils. In practice, however, incumbents suspected that parent councils would rubber-stamp the decisions made by principals, who remained government-appointed and were not affiliated with unions.^[Interviews ES4, ES7, ES8]   This indeed came to happen. The relationship between school principals and rural parents (some unable to read) was unequal, and principals thus maintained substantial authority in their communities. In one of my interviews a former EDUCO teacher admitted that he had been hired on the recommendation of a local ARENA party leader who had contacted the school principal.^[Interviews ES9]   While it is impossible to know how common this practice was, this anecdote showcases that at different levels of government, the party’s commitment to EDUCO as a tool to improve educational outcomes was limited.


The political implications of decentralization were straightforward. By replacing *escuelas populares* with official schools, and staffing them with teachers unaffiliated with ANDES, the government would exclude teachers’ unions from rural areas. Instead, ARENA filled jobs with its own supporters, planting its own political brokers. It also ensured that the new teachers would never coherently organize against them. Given that contract teachers faced different employers, organizing was time consuming and costly. My informants confirmed that any whiff of unionization was swiftly met with termination.^[Interview ES5, ES8] The government pursued EDUCO far beyond the original FMLN territories, expanding it to urban areas in the late 1990s and eventually covering 40% of all children in basic education [@meza_educo:_2004].  In short, decentralization effectively hobbled ANDES activity by weakening its presence across the country.

A related question remains: why was the government successful in pursuing EDUCO? This account explains the logic that motivated ARENA to pursue decentralization fervently. But this alone does not explain why it succeeded. A review of the facts suggests that the government had help from circumstantial factors. On the one hand, Gallardo was clever in designing the reform, both by developing a program (EDUCO) with roots in rebel organizations and by leveraging the support of the international community and prominent local actors.^[Interviews ES1, ES2, ES3]  The fact that EDUCO schools were based on *escuelas populares* was meant to serve as an olive branch to left-wing actors. But as the political narrative described above reveals, it actually fully gutted *escuelas populares* and replaced opposition actors with those more favorable to the government. The government easily won the support of international organizations, development banks, and other education reformers by selling this administrative decentralization program on its hypothesized technical merits. It was “sold” as a cost-effective, democratizing way of improving the coverage and quality of education. The mass of support for the program from actors other than the opposition made it difficult to challenge and established the origins of EDUCO as *technical* rather than *political.*

On the other hand, most political actors viewed the transition period as tenuous, and therefore acted cautiously. Interviewees on the left and the right confirmed that the general sense among the population was one of exhaustion with social conflict.^[Interviews ES11, ES14]  Thus, while ANDES mobilized large protests against education decentralization reforms on a few occasions, including a more severe decentralization program modeled after the Chilean experience,^[Interviews ES1, ES3]   it eventually found itself bargaining with the government over the program. What did ANDES leaders extract from the government? The evidence is unclear, but some of my interviewees suggested that Gallardo promised ANDES leaders that most urban teachers (their electoral base) would remain excluded from the EDUCO system.^[Interviews ES5, ES11]   Division within the FMLN was also significant. Some officers believed that an FMLN win would reverse democratization, splintering the organization.^[Interviews ES14]  Thus, the FMLN was in a weak position to defend ANDES. In short, the specific dynamics of the transition and post-conflict period meant two things: 1) incumbents were relentless in their efforts to decentralize education while 2) ANDES and the FMLN were in no position to fully prevent it. The result was education governance decentralization.

The Salvadoran case study shows the political-electoral origins of education reform. The presence of a tie between teachers and the opposition incentivized ARENA to advance a very specific decentralization program meant to weaken the organizational capabilities of teachers, and therefore the opposition. EDUCO cannot be explained simply as a strategy to expand public education to war-torn regions; the logic of its expansion and its operation in practice reveals a very carefully targeted program. An unresolved question is whether education leaders, aside from their political motivations, actually believed that the program would improve the coverage and quality of education. My argument is entirely consistent with this possibility: even if they thought EDUCO would improve education outcomes, it is improbable that they would not have seen an electoral benefit to excluding ANDES-affiliated teachers from the countryside. Indeed, in the absence of such an incentive, reform would not have been advanced.


## Paraguay: No Decentralization Proposals in the Absence of Union-Opposition Ties

The history of public education governance in Paraguay is like that of El Salvador. Starting in 1933 the national government made a concerted effort to control education directly, generating a fairly centralized education system that remained untouched for most of the century. By the end of the 1980s this education system had delivered poor results, especially compared to other Latin American countries. Reports commissioned by the @world_bank_paraguay_2002  pointed to low completion and achievement levels among students, particularly those from rural communities and poor backgrounds. In 1990 primary completion rates sat at 65%, roughly on par with El Salvador, and were lower still for the rural student population, which accounted for half of all students [@world_bank_world_2018].  In sum, there was a technical case to be made for the improvement of public education.

The overarching concern of the conservative party, the *Colorados*, however, was to remain in power under democracy, not education reform. General Stroessner was toppled in 1989 by his chosen successor, General Andrés Rodríguez, who subsequently initiated a transition to democracy. Democracy brought new challenges for incumbents. The first sign of trouble emerged in 1991 when Carlos Filizzola of the *Asunción Para Todos* party was elected mayor^[Technically, Filizzola was elected *intendente municipal*.]  of Asunción despite trailing the *Colorado* candidate, Juan Manuel Morales, in official polling.^["Paraguay celebra sus primeras elecciones municipales directas." El País, 26 of May, 1991, accessed on March 15, 2020 available: [here][link2].]
Filizzola’s victory offered a warning: in the post-transition period the Colorados could not assume they would triumph at the ballot box [@velazquez_seiferheld_relaciones_2018]..  In the run-up to 1993 this was confirmed by official polling, which placed the *Colorado* candidate, Juan Carlos Wasmosy, in an embarrassing third place.

[link2]: https://elpais.com/diario/1991/05/27/internacional/675295206_850215.html


Counter to the expectations of the technical and ideational arguments, education reform under the right-wing *Colorados* was slow-moving. Recognizing that *something* had to be done, incumbents took a hands-off approach. In 1990 President Rodríguez established an education reform council, *Consejo Asesor de la Reforma Educativa* (CARE), and charged it with charting the path forward on education. To the relief of reformers, council members were well-respected independent experts who exhibited relative autonomy from the government in their deliberations [@nunez_reforma_2002]. CARE considered and debated many different types of education reforms, influenced in part by prominent ideas among reform advocates, international organizations, and the Harvard Institute for International Development [@rodolfo_alisis_2014].  Decentralization was thought to be a pillar of education reform.  Indeed, CARE’s first proposal, in 1992, centered reform around three objectives: 1) administrative changes in the Ministry of Education, 2) the selection of teachers, and 3) decentralization.^["Reforma Educativa, Compromisode Todos" 1992.]  

Incumbents, however, never seriously entertained decentralization. The reason is that teachers were not allied with the opposition; in fact, for most teachers, the reverse was true. During the Stroessner era, public sector jobs were tightly controlled by the dictatorship. To be eligible for a teaching position, for instance, individuals needed to be *Colorado* party members and required a written recommendation from a mid-tier party or military leader [@rivarola_reforma_2000]. The largest teachers’ union, the *Federación de Educadores del Paraguay* (FEP), with roots in dictatorship, was thus aligned with the Colorados [@lopez_sindicatos_2009; @becker_sindicatos_2009].  The practice of appointing teachers on the basis of their political affiliation continued well into the 2000s, when the *Colorados* were voted out of office in 2008 with the election of Fernando Lugo. Institutional constraints, including tenure protections, made dismissals of civil servant posts unheard of and thus extremely valuable [@schuster_when_2016]. The government doled out these positions carefully. Predictably, FEP mobilization was an important part of the *Colorado* electoral strategy. In the run-up to the 1993 elections, teachers were mobilized in rural areas to get out the vote in favor of the *Colorados*, leading to a high-profile clash with the Minister of Education, Raúl Sapena Brugada, who resigned rather than oversee a partisan education system [@nickson_wasmosy_1997; @harding_paraguay_1993]. 

From a political perspective, the *Colorados* might have reaped an electoral reward from decentralization by weakening the teacher labor movement more generally. Democratization meant that new unions could form and participate in politics. Other, smaller teachers’ unions active around this time included the *Organización de Trabajadores de la Educación del Paraguay-Sindicato Nacional* and the *Unión Nacional de Educadores* [@becker_sindicatos_2009].  These organizations had a contentious relationship with the incumbents, deploying strikes to bargain over the improvement of teacher working conditions. Protests led by these organizations in 1990 and 1993 culminated in significant wins for teachers, including a flat wage increase of 30% [@nunez_reforma_2002].  These unions (alongside FEP) benefited from collective bargaining with the central government because they could credibly strike, bringing education across the country to a sharp stop. If education decentralization hurts teacher solidarity, pursuing decentralization as CARE suggested could have weakened unions’ ability to mobilize for their interests. However, it also might have implied transferring the responsibility of appointing or monitoring teachers to subnational governments. Given the alarm-raising successes of the opposition in subnational elections, national incumbents were understandably wary of these projects.

Efforts to decentralize education therefore fizzled, despite the prominence of CARE and a larger push to grant subnational units (*departamentos* and *municipios*) political and fiscal authority in the 1992 constitution. For example, the 1993 *Ley Orgánica Municipal* provided that municipalities *could* run their own schools if they so wished, but apportioned no national funds for such activity. Indeed, with the exceptions of Asunción and San Lorenzo, there are no municipal schools in Paraguay [@gaete_estudio_2012]. At the department level, the 1998 *Ley General de Educación* mandated that the Ministry establish a subnational council to coordinate efforts with the *departamentos.* But these councils are at best weak advisory bodies to the Ministry; they do not hire teachers, set education policy, or administer or manage schools [@letelier_s_education_2018].  

In sum, education in Paraguay by the end of the 1990s can only be characterized as strongly centralized, if with some allowances for subnational units to convey concerns over education to national authorities. Looking back on the promise of the education reform movement, the perception among many reform advocates is that things did not change radically. In the words of a teacher not affiliated with FEP: “they have engaged in curriculum reforms, that theoretically and pedagogically where OK..but everything pretty much remains as before” [@palacios_de_asta_descentralizacion_2002].  This uneven commitment to decentralization by the *Colorados*—who remained in power until 2008 and thus had plenty of further opportunities to pursue reform—can only be explained by the absence of an opposition-union link.


# Implications for Existing Work and Future Research Avenues 

The paper shows that alliances between teachers’ unions and the opposition can compel incumbents to advance decentralization with the deliberate intent of weakening teacher organization. This theory is utilized to properly contextualize the political incentives that generate support from incumbent politicians for decentralizing reforms.

```{r setupfig2, include=FALSE}
## Bayesian Regressions

## Load Packages
library(tidyverse)
library(socviz)
library(hrbrthemes)

## Read in data
pol<-read.csv("data/pres_political_data.csv") %>%
  mutate(year = as.factor(election_year),
         rates = completion_rate/10,
         oul = teachers_with_opp)

mini<-pol %>%
  select(education_decentralization_reform, completion_rate,
                        right_center_of_gravity, imf_education_loans, teachers_with_opp, 
                        polity, oul)
#stargazer(mini) # For appendix

data<-read.csv("data/presidential_admin.csv")
fig2<-ggplot(data, aes(x=election_year, y=decentralization_level,
                       linetype=as.factor(teachers_with_opp),
                       color= as.factor(teachers_with_opp),
                       shape=as.factor(teachers_with_opp))) +
  geom_jitter(size=2) +
  geom_smooth(method = lm, se= F, size=3) +
  labs(x="Election Year", y="Decentralization Level", linetype="Opp-Teach", color="Opp-Teach") +
              theme_bw()+
            theme(panel.border = element_rect(fill = NA, size = 1),
                  legend.position = "none") +
              coord_cartesian(xlim=c(1980,2010), ylim = c(0,1))+
            labs(y="Level",
                 x="Year",
                 title = "Education Decentralization",
                 subtitle = "Across Teachers' Union Opposition Ties") 

fig2a<-ggplot(data, aes(x=election_year, y=decentralization_level,
                       linetype=as.factor(teachers_with_opp),
                       color= as.factor(teachers_with_opp),
                       shape=as.factor(teachers_with_opp))) +
  geom_jitter(size=2) +
  geom_smooth(method = lm, se= F, size=3) +
  labs(x="Election Year", y="Decentralization Level", linetype="Opp-Teach", color="Opp-Teach") +
              theme_bw()+
            theme(panel.border = element_rect(fill = NA, size = 1),
                  legend.position = "none") +
              coord_cartesian(xlim=c(1980,2010), ylim = c(0,1))+
            labs(y="Level of Decentralization",
                 x="Year") 

ggsave(fig2a, file="figures/fig2.pdf", width = 4, height = 3)

```

```{r, plotfig2, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, fig.cap = "Association Between Opposition-Union Links and Decentralization", fig.width=4, fig.height=3, cache= F, label=fig4}
fig2
```

This paper makes several contributions to the literature on decentralization, parties, and education politics. First, it identifies a new motivation for decentralization. This piece is the first to show that incumbents do so to kneecap their political opponents. This understanding of the reasons why politicians pursue or enact decentralization policies undercuts a common framing in the literature. Strategic politicians—not powerless politicians desperate for development bank cash or neoliberal ideologues—advocated for structural changes at least in part because they provided them (or so they thought) with a concrete electoral benefit.^[Thanks to an anonymous reviewer for this point.]  Further work needs to untangle how different aspects of much-touted decentralization projects are pursued specifically for these types of feedback effects.

Second, while my argument cannot fully explain why governments eventually succeed or fail at implementing their decentralization reforms, it does have important implications for this line of research. If incumbents are at least partially motivated by hidden political logics when advancing decentralization, then they are very likely to persist in pursuing them when the political conditions are right. I show this for El Salvador. But can this explain other cases? Consider the following plausibility test.^[Details on data are in the Appendix.]  Figure 2 plots the level of education decentralization for presidential administrations across eighteen Latin American countries from 1980 to 2010. I sort presidential administrations into two groups: those in which I can identify a linkage between any teachers’ union and an opposition party and those in which I cannot. Those with ties are represented by triangles; those without are represented by dots. I also overlaid regression lines for each group. The trend for administrations with union ties is in a dashed line; the trend for administration without is in a solid line. The figure clearly shows that over time, education decentralization was more likely to be successfully enacted when the opponents were aligned with teachers. Additional research is needed to test this intriguing finding.

Third, my work adds to our understanding of how political parties operate to win elections in Latin America. Much of the literature on this topic focuses on clientelism and vote-buying [@nichter_vote_2008;@bustikova_patronage_2017;@stokes_brokers_2013;@szwarcberg_uncertainty_2012].  Less attention has been given to the ways parties advance (legal), but hidden, policies to limit their opponents’ ability to reap electoral rewards. In recent work Niedzwiecki -@niedzwiecki_uneven_2018 has shown that partisan alignment between subnational and national governments shapes how and when social policies are well implemented. The reason is that subnational politicians anticipate political rewards to social policies and thus are wary of advancing policies that will benefit their opponents. This contribution shares with my argument the idea that parties understand the resources that their opponents draw on to win elections, and that policies have feedback effects. More work is needed to assess how parties promote or slow-walk certain policies with the intention of hurting their competitors.

Finally, my research advances the literature on education politics. A robust literature on education reform emphasizes the role that teachers’ unions play in blocking reforms [@corrales_politics_1999; @moe_comparative_2016; @diaz_rios_when_2016].  In this scholarship the partisan attachments of teachers’ unions tend to protect them from “bad policies” [@murillo_recovering_1999].  I push back against the idea that parties or incumbents want to break up unions because they are an obstacle to improving the quality of education. Instead, I raise the important point that sometimes, perhaps more often than not, teachers are casualties of policies designed to limit their mobilizational advantages *because* they have partisan ties to opposition groups. The idea that partisan linkages are a double-edged sword should garner more attention in this literature.

# Bibliography