---
title: 'The Political Origins of Education Decentralization'
author: "Thomas R. Vargas^[Postdoctoral Fellow, Department of Political Science, McMaster University, Hamilton, ON. Corresponding address: vargasct@mcmaster.ca. I thank Teri Caraway, Michelle Dion, John Freeman, Claire Le Barbenchon, Dave Lopez, Oanh Nguyen,  Oscar Pocasange, David Samuels, Ben Ross Schneider, and two anonymous reviewers for their comments on previous drafts of this paper. I am also grateful to seminar participants at the University of Minnesota and Duke University, and audiences at APSA 2019 and MPSA 2018. My thanks to the Department of Political Science at the University of  Minnesota for research support and to the World Bank Archives for access to key material. Any remaining mistakes are my own.]"
abstract: \singlespacing What explains why some countries advance decentralized education while others shy away from such efforts? In the 1990s several Latin American governments reformed their education systems to give local communities more control over the management of public education. Some arguments suggest that decentralization was pursued to improve the coverage and quality of education (Grindle 2004; Kaufman and Nelson 2004). Others point to partisanship, ideology, or diffusion. Drawing on an in-depth case study of El Salvador with reference to Paraguay, I show that education reform in Latin America was driven in part by political processes that had little to do with any of these explanations. I argue that education decentralization was pursued because it could be deployed as a political weapon. Specifically, I show that the incumbent political party supported education decentralization because they intended to weaken teachers’ unions affiliated with the opposition, thus depressing mobilization and votes for their rivals. These findings contribute to the literature on decentralization by highlighting a new political motivation fueling decentralization efforts across the developing world--the demobilization of the opposition. It also contributes to the literature on education by showing how the associational ties between teachers and parties, often thought to protect them, can actually set them up to lose.
date: "This draft: July 20, 2021"
bibliography: setup/bib.bib
csl: setup/comparative-politics.csl
mainfont: Times New Roman
sansfont: Times New Roman
colorlinks: true
documentclass: article
geometry: margin=1in
fontsize: 12pt
linestretch: 1.5
link-citations: true
output:
  pdf_document: 
        includes:
          in_header: setup/header.tex
        number_sections: true
  html_document: default
  word_document: 
    reference_docx: setup/word-template.docx

---
\newpage

# Introduction

Public investments in primary education have huge payoffs in terms of long-term development outcomes. Many countries in the developing world, however, struggle to improve the quality and coverage of public education. One widely touted reform designed to improve such outcomes is *decentralization.* In this regard, scholars are presented with a puzzle: some countries around the world with low educational achievement and coverage rates advanced the decentralization of primary and secondary education, while others with similar attributes did not. If---as prominent observers have suggested [@grindle_despite_2004; @kaufman_politics_2004]---governments pursue decentralization in order to improve access to education (a “technical” goal), the explanation for such variation remains a mystery. I argue, by contrast, that governments often pursue a decentralized provision of education for explicitly political purposes. Focusing on explaining variation across Latin America specifically, I claim that decentralization can strengthen the electoral position of incumbent political parties by weakening the opposition whenever it is allied with teachers’ unions.

Consider the cases of El Salvador and Paraguay. In 1991 the Salvadoran government pushed for a school-based management program called *Educación con Participación de la Comunidad* (EDUCO) that would eventually devolve the governance and administration of most rural schools to local communities. Under EDUCO, rural communities would oversee infrastructure, collect school fees, and hire teachers—a truly extensive and radical rethinking of how to manage the provision of public education in the region. By contrast, Paraguayan authorities did not push for the decentralization of education despite their country's many similarities to El Salvador during its EDUCO years: it had low primary education completion rates, it was beset by high levels of poverty in rural areas, and it was ruled by a right-wing government, (for other similarities, see table 1). While Paraguayan elites were aware of decentralization programs, governance reforms in education were muted.

The difference in support for decentralization projects can be explained by looking at how incumbents viewed the partisan affiliations of teachers. Following the Salvadoran civil war (1980–1992), parties faced considerable uncertainty regarding their electoral prospects under democracy. The ruling right-wing party, *Alianza Republicana Nacionalista* (ARENA), feared losing power to a guerrilla-turned-legitimate political actor, the *Frente Farabundo Martí para la Liberación Nacional* (FMLN). To avert the electoral success of the left, ARENA sought to limit the ability of teachers’ unions to mobilize in favor of the FMLN, their political allies, by pursuing education decentralization. In Paraguay the absence of a substantive link between teachers’ unions and the opposition party meant there were no political incentives to reform the governance of education, even though the education system’s performance (measured by completion rates, for instance) was as weak as El Salvador’s.

These examples showcase the political logic that can help explain why some Latin American governments pursued reforms to education governance in the 1990s while others did not: incumbent political parties supported decentralizing reforms because they weakened teachers’ unions affiliated with opposition parties. In turn, they expected this to result in fewer votes for the opposition at the ballot box. This argument is developed by bringing together the extant literature on the electoral advantages of political and fiscal decentralization for incumbents [@montero_decentralization_2004;@oneill_decentralization_2003;@oneill_decentralizing_2005; @niedzwiecki_social_2016] with scholarship exploring the partisan identities of teachers' unions [@larreguy_political_2017;@moe_comparative_2016; @murillo_recovering_1999]. I also draw from work documenting the use of policy as a political weapon originally developed in the context of partisan competition in the United States [@anzia_politicians_2016; @hertel-fernandez_policy_2018]. 

This study contributes to our understanding of decentralization politics in two ways. First, I document a new political motivation for decentralization. Previous work by O'Neill -@oneill_decentralizing_2005, Dickovick -@dickovick_municipalization_2007, and González -@gonzalez_political_2008 has shown that incumbent parties undertake decentralization when their support at the national level is waning. The logic is that national incumbents are more competitive in the newly empowered subnational units. In the argument developed here, political incumbents do not withdraw to newly created competitive environments, they *reshape* the existing (national) arena to their advantage by targeting opponents and their allies. This theory draws on work by Hertel-Fernandez -@hertel-fernandez_policy_2018 in the American states showing that conservative groups advance policies with the deliberate intention of demobilizing political opponents. This same underlying political logic can help explain why seemingly nonpartisan education decentralization efforts were pursued with such fervor in certain contexts and ignored in others.

Second, this paper challenges the idea that decentralization can be thought of as a wholesale process entailing the concurrent devolution of administrative, fiscal, and political responsibilities  [@falleti_decentralization_2010]. In Falleti's -@falleti_decentralization_2010 classical framework, the decentralization of education (which she calls an "administrative" reform) cannot be separated from devolution projects in other domains (fiscal and political). In examining the case of El Salvador, I establish that administrative decentralization with no corresponding reforms shaping political or fiscal authority is possible, that it is not toothless (i.e. it can utterly reshape the national provision of public education), and that it can accomplish significant political goals. This finding builds on work by @bohlken_democratization_2016, who notes that incumbents in India advance political decentralization without concurrent fiscal or administrative decentralization to further their own political aims. Because administrative changes are couched in "technical" rather than political arguments, they are possibly easier for governments to push for. Disaggregating wholesale decentralization projects into component parts and examining their underlying political motivations is thus important since it can  open up intriguing theoretical possibilities to scholars of policy and decentralization.

Finally, this study pushes forward the literature on education politics. Most of this scholarship emphasizes how the process of education decentralization unfolds, assuming away the reasons incumbents push for decentralization and focusing instead on the strategies teachers' unions use to limit reform [@grindle_despite_2004;@kaufman_politics_2004;@corrales_politics_1999; @murillo_recovering_1999; @falleti_decentralization_2010]. My dependent variable is not the ultimate success or failure of decentralization efforts, but whether or not incumbents select decentralization as a policy in the first place. In doing so I shift focus slightly away from teachers' unions. Still, I do not dispute the dominant paradigm: teachers can be powerful players in the politics of education, and their partisan attachments inform the strategies they follow in defending their vested interests [@moe_comparative_2016; @bruns_politics_2019].  But an emphasis on what teachers *do* (specifically, for instance, block reform) understates a central fact that structures democratic politics everywhere: parties want to win elections. Therefore, we should not assume that governments propose politically costly changes to education systems exclusively because they want to improve them. In the cases at hand, I show that politicians take advantage of decentralization policies whose second-order effects demobilize their rivals (or so they believe.) Viewing education reform through this electoral lens can turn received wisdom on its head: the partisan attachments of teachers’ unions, broadly believed to empower them to resist change [@murillo_recovering_1999], can actually set them up to lose.

# Existing Accounts of Decentralization Motivations

The earliest work on education decentralization in Latin America suggests that the origins of education decentralization spring from one of three causes: technical factors, ideational pressures, and diffusion. As I show below, each is insufficient to address the puzzle at the beginning of this paper.

## The Technical Explanation

The technical determinants theory is the most prominent among scholars of education reform. It holds that some governments in the region decentralized because their public education systems were overwhelmed by a crisis of coverage and quality [@grindle_despite_2004;@kaufman_crucial_2004]. Through the 20th century, reformers argued, Latin American governments created bureaucratic behemoths that were too slow to meet these challenges. According to this line of thinking, the only viable solution to improving the coverage and quality of education was to break up concentrated authority. This would achieve gains in efficiency at the same time that it democratized decision-making by devolving decisions over education to local communities. In countries with lagging enrollments in far-away rural areas, like Paraguay and El Salvador, decentralization would would make it easier to expand schooling without having to go through the national bureaucracy. 

```{r, figure1, include=FALSE}
## Load packages
library(tidyverse)

## Load data
qog<-read_csv("data/qog_std_ts_jan19.csv") %>% # QoG dataset
  select("cname", "year", "ccodecow", "ccodealp", # select only relevant variables
                           "wdi_expedu", "bmr_dem", "p_polity2", "wdi_pop14",
                           "gle_cgdpc","gle_trade", "wdi_expeduge", "wdi_expedup",
                           "wdi_expedus", "ht_region",
                           "wdi_expedut", "wdi_expstup", "wdi_expstus", "wdi_expstut",
                           "bl_lpmf" , "wdi_popurb", "wdi_debt", "bl_asymf", "wdi_gerp",
                           "wdi_nerp") %>%
  rename("cowcode" = "ccodecow" ) 

## Merge data 

data<-read.csv("data/TRV_decentralization.csv") %>% # Decentralization data
  left_join(qog, by=c("cowcode", "year")) # Merge with QoG

## Plot performance data

data%>%
  select(full_name, year, wdi_gerp) %>%
  arrange(full_name)%>%
  subset(year>1977)

f1a<-data %>%
  ggplot(mapping=aes(x=year, y=wdi_gerp)) +
        geom_smooth(color="black", size=2) +
        geom_jitter(alpha = 0.2) +
        theme_bw() +
        xlab("Year") +
        ylab("Enrollment Rates") +
        coord_cartesian(xlim = c(1970,2010), ylim = c(0,120), expand = TRUE,
                        default = FALSE, clip = "on") +
        scale_x_continuous(breaks = seq(1970, 2010, by = 10)) 

f1b<-data %>%
  ggplot(mapping=aes(x=year, y=wdi_nerp)) +
        geom_smooth(color="black", size=2) +
        geom_jitter(alpha = 0.2) +
        theme_bw() +
        xlab("Year") +
        ylab("Completion Rates") +
        coord_cartesian(xlim = c(1970,2010),ylim = c(0,120), expand = TRUE,
                        default = FALSE, clip = "on") +
        scale_x_continuous(breaks = seq(1970, 2010, by = 10))

library(gridExtra)
```

```{r, plotfig1, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, fig.cap = "Enrollment and Completion Rates for School-Aged Children Latin America Since 1970", fig.width=5, fig.height=3, cache= F, label=fig1}
grid.arrange(f1a,f1b, ncol=2)
```

One problem with this argument is that the data available at the time governments began decentralizing was mixed and limited. In terms of coverage, most reports from the time document drops in enrollment in secondary schooling. But, as shown in Figure 1, the untold story of the 1980s is that the region experienced no reversals in primary school coverage -- a remarkable achievement. Problems were more intuitive in terms of quality. In poor and under-resourced areas, attending school does not guarantee learning. While it reasonable to believe that education quality was generally low across the region, empirical evidence regarding the quality of education was nonexistent until well into the 2000s. As documented by the UNESCO in 2001: "we do not yet have a time series of education achievement indicators that is sufficiently broad to adequately describe the behavior and rends of such indicators within countries in the region" [@unesco_state_2001, p.10]. That Latin American governments supposedly concerned with learning didn't collect more evidence regarding their quality-improve reforms is highly suspect.

Still, it is plausible that regional trends in decentralization were driven by countries that, on average, trailed their peers. For instance, countries with the lowest enrollment rates (which in the absence of learning data might reasonably proxy for quality issues also) may have been under the most pressure to decentralize. This would be consistent with the notion that technical considerations propelled reform. As the cases of El Salvador and Paraguay illustrate, however, there are discrepancies even within countries with "weak" education systems. In short, to account for regional variation in the decentralization of public education, we have to look beyond technical concerns regarding the coverage and quality of schools.  

## Ideational Theories 

The ideational theory of education reform attributes decentralization to neoliberal policy ideas that gained traction among center-right and right-wing parties. These parties, coming to power in the 1990s, were keen on introducing market-oriented strategies to the provision of social welfare. In the face of pervasive deficits, these governments were interested in keeping social expenditures sustainable and social policy programs cost-effective [@nelson_social_2011].  We would therefore expect right-wing governments to decentralize education, whereas center or left-wing governments would shy away from such efforts.  

The neoliberal logics of decentralization are easy to outline. For states burdened with debt, passing responsibility for the management and funding of education to subnational units would reduce central government spending [@falleti_decentralization_2010]. In theory, aggregate education expenditures would also fall, because the cost associated with collecting knowledge about prices and production processes is lowest at the local level [@shah_balance_1998]. Similarly, decentralization was thought to yield innovation in the provision of services as long as there was enough delegation of responsibility [@oates_fiscal_1972; @ford_rationale_2000; @tiebout_pure_1956].  For neoliberal advocates, decentralization offered the potential efficiencies and innovation associated with free markets. Yet decentralization has not been consistently associated with a single political-economic logic [@montero_decentralization_2004]. During the time under analysis, both right- and left-wing governments proposed education decentralization programs. Perhaps most interestingly, various right-wing governments have shied away from decentralization. For example subsequent administration of Paraguays' *Colorados* never seriously considered decentralization. 

It's plausible, nonetheless, that it was those states most burdened by the debt crisis that were compelled to decentralize education as a cost-saving mechanism. This is another variation of the technical incentives argument that government officials used in trying to sell decentralization. But countries saddled with high levels of debt in the aftermath of the debt crisis followed different trajectories with regard to decentralization. Take again the examples reviewed at the opening of this paper. While both El Salvador and Paraguay experienced sharp increases in debt service as a share of GDP during the 1980s, by 1989 Paraguay's outstanding external debt accounted for .53 of GDP compared to .50 for El Salvador [@huber_social_2012]. Yet, it was El Salvador and not Paraguay engaged in any decentralization efforts. In short, fiscal considerations alone cannot explain education reform. 

## Diffusion Explanations

Finally, the diffusion arguments suggest that governments began making certain types of changes to their education systems to accommodate both international financial institutions (IFIs) and changing global norms surrounding schooling. The early 1990s represented a unique time in international education. Donors and international organizations heavily promoted the idea that education was crucial to development. This is best exemplified by the World Conference on Education for All held in Thailand in 1990—financed by international organizations—which set five target education goals for 155 countries. One possible mechanism is that these organizations, including the World Bank and the IMF, pressured governments into adopting their own ideas about “what works” in education by funding certain education reform projects, or by attaching conditionalities to structural adjustment programs (SAPs). We would therefore expect that countries with active SAPs or sectoral loans with IFIs would be more likely than others to decentralize education.

To assess whether this was plausible, I examined 193 IMF loan agreement documents, including letters of intent, arrangement letters, and staff reports, for 18 Latin American countries from 1980 to 2000.^[Thanks to XX for sharing her collection of documents with me.] I looked for conditions that required education sector reform (of any type) prior to the disbursement of loans -- what are known as "hard" conditions [@woods_globalizers_2014]. I found no trace of "hard" conditions placed on the education sector. I did uncover two instances of "soft" conditions, conditions used mostly to track progress in program implementation [@caraway_international_2012;@kentikelenis_imf_2016], in Argentina and Bolivia. These soft conditions, however, were meant to encourage the implementation of decentralization programs already approved by the respective legislature; not the advancement of new decentralization projects. In sum, there is limited evidence suggesting that IFIs coerced or enticed governments into pursuing decentralization.

On the other hand, international organizations surely helped shape the conversation around education reform in the 1990s [@edwards_jr_trajectory_2017;@mundy_world_2016]. IFIs may have relied on their convening power to instill on politicians and high level bureaucrats the importance of decentralization. While this is plausible, conventional wisdom tends to overstate how emphatically agencies like the World Bank advocated for education decentralization specifically. For example, in working papers widely circulated around the Bank in the 1980s and the mid 1990s, Bank staff warned that the decentralization of public services was no panacea, especially in poor countries with weak states [@rondinelli_decentralization_1983; @prudhomme_dangers_1995]. My exploration of recently declassified internal World Bank correspondence related to various education reform efforts in El Salvador and Paraguay suggests that the Bank was more interested in increasing expenditures in primary education than in advocating for any particular education governance reform. It remains possible that individual IFI staff advocated heavily for decentralization convincing mid-level bureaucrats and some politicians that governance reforms were necessary to fix some quality or coverage issue. But this does not explain *why* governments finally decided to expend political capital on pursuing a costly endeavor. Thus, while ideas about decentralization were surely available and known to bureaucrats and politicians, this was not the principal *driver* of reform.

# Decentralization as a Political Weapon to Demobilize the Opposition

My argument is that the partisan affiliation of teachers determines whether or not incumbents push for education decentralization. The reason is that decentralization strengthens the electoral position of incumbent political parties whenever their opponents are affiliated with teachers’ unions because decentralization tends to weaken teachers’ organizational advantages, which are often used to help their partisan allies drum up votes.

Political scientists have long noted that policies shape politics in intended and unintended ways. I focus here on the notion that politicians "make politics" by advancing legislation and policies aimed at creating feedback processes that benefit those who propose them politically [@hertel-fernandez_policy_2018; @anzia_interest_2019]. In my argument, incumbents understand that education decentralization has first- and second-order consequences. The first-order consequences refer to the specifics of education reform: for example, changes to who hires or monitors teachers. These changes are the ones politicians "sell" by relying on technical arguments. Pointing to gaps in quality and coverage, incumbents propose and pursue these decentralization arrangements. But they are deeply interested in the policy's knock-on effects. In certain cases, the decentralization of education can weaken teacher's unions. To the degree that the opposition benefits from an associational tie with these unions, decentralization can weaken the opposition's mobilization advantages. Hertel-Fernandez's calls the strategy of demobilizing the opposition via policy feedback effects *policy as political weapon*. 

The idea that decentralization "makes" politics has been documented before.  @oneill_decentralizing_2005, for instance, has argued that incumbent political parties give up centralized fiscal authority to subnational units whenever two conditions are met: they fear losing power at the national level and they believe their electoral fortunes are more secure at lower levels of government. More recently, @bohlken_democratization_2016 has shown that higher level government elites use local political decentralization reforms as a means of exerting control over lower level intermediaries.  By introducing local municipal elections with no corresponding fiscal autonomy, subnational elites solve two problems: they can more clearly identify effective intermediaries (i.e. people that win elections) at the same time that they make them increasingly dependent on the subnational party for fiscal resources. In all cases, incumbents choose decentralization because, counter-intuitively, it strengthens their electoral or political position.

I borrow the concept of policy as weapon to describe the political motivation behind education decentralization. Whenever incumebnts know that the opposition relies on teachers to mobilize voters, they advance decentralization to weaken their ability to help their partisan allies. To my knowledge, this is the first empirical account to show decentralization as a weapon to demobilize the opposition. 

I build this argument from a series of insights about the behavior between incumbents, opposition parties, and teachers. To begin, we must recognize that in the aftermath of the Third Wave national elections are the most important way for parties to access political power in Latin America [@mainwaring_juan_1997]. If incumbents lose, their ability to shape policy and reap rewards for themselves and allies diminishes. To the degree that policies determine electoral outcomes, successful incumbents will select policies so as to ensure their survival [@downs_economic_1957; @riker_theory_1984]. Incumbents and parties are thus strongly motivated by elections [@mayhew_congress_1974;@samuels_ambition_2003]. Under this view of the world, governments must approach costly, highly visible changes to social policy—like the decentralization of education—at least partly through the lens of re-election [@whitfield_what_2011]. Politicians advancing education decentralization, therefore, must have good political incentives to do so. 

A second point is that in Latin America, as elsewhere in the developing world, parties rely on brokers both to collect information on voters and deliver votes [@stokes_brokers_2013; @szwarcberg_uncertainty_2012; @lowden_education_2004; @eaton_teachers_2014]. While scholars have previously pointed out how state employees can play the role of brokers [@weitzshapiro_what_2012; @oliveros_making_2016; @mares_core_2018], teachers are uniquely positioned to do so because of their distribution, numbers, and position of authority. Anecdotally, the notion that teachers’ unions have deep and steady partisan affiliations and deliver votes for their parties is common across the region.  @lowden_education_2004 reports that in Colombia it was typical for teachers to be hired *en masse* around election time. In Mexico, many teachers affiliated with the Mexican National Educational Workers' Union (SNTE) similarly owed their jobs to PRI leaders, who weaved a web of patronage-based practices regarding the hiring and management of teachers [@chambers-ju_teachers_2017]. Furthermore,  @larreguy_political_2017 have shown empirically that the PRI’s electoral victories were secured in part by the mobilizing efforts of the SNTE and that these attachments persist. Indeed, across much of the developing world, there is mounting evidence that teachers act as political brokers and that their unions are caught up in complex webs of clientelism and patronage [@sandholtz_voters_2019;@beteille_absenteeism_2009;@chattharakul_thai_2010;@pierskalla_personnel_2019;@eaton_teachers_2014]. What is analytically key is the importance of actors in this sector to political parties. This sets education politics apart from other types of sectoral decentralization, such as health. 

A third insight is that the costs of education decentralization fall disproportionately on teachers. Teachers have organizational and mobilizational advantages over most public employees. Not only do teachers typically outnumber every other public employee group by a fair margin[@bruns_great_2015], but they are also typically dispersed across all areas of a country and are deeply integrated into local communities [@moe_comparative_2016]. These advantages are compounded by centralized bargaining because teachers can credibly threaten to paralyze a country when a government fails to meet their demands. When education systems shift to decentralized governance, however, these advantages weaken because a single teachers’ union now must contend with multiple employers. More specifically, as a result of decentralization, teachers lose cohesion—because teachers face differing hiring and incentive structures across a national territory—and therefore negotiating leverage, at least with the central government [@olson_logic_1971]. In turn, teachers expect to face lower wages, less access to jobs, and worsening working conditions [@hanson_educational_1997; @murillo_recovering_1999]. For these reasons, teachers’ unions tend to strongly oppose decentralization [@grindle_despite_2004; @kaufman_crucial_2004], sometimes to the point of violence [@bruns_managing_2016]. 

A third insight is that relatively stable party-teachers’ unions linkage establishes parties as an arm of either the government or the opposition. If teachers’ unions’ partisan affiliations are relatively slow moving, then incumbents not aligned with teachers have an incentive to disrupt teacher organization. 

My argument doesn't require that the links between parties and teachers be permanent, only that incumbents view them as relatively stable.  If teachers’ unions were free to quickly switch allegiances to the party in power, they may be able to better resist decentralization efforts and gain other benefits. On the one hand, teachers’ unions must maintain some degree of ideological coherence in their partisan attachments. This means that they are ideologically constrained in their alliances. Teachers’ unions that espouse communist ideals, for example, cannot easily switch sides and support conservative, right-wing parties. In short, the ideological content of unions should be relatively sticky. Still, as @chambers-ju_adjustment_2021 has pointed out, in the aftermath of decentralization these links underwent considerable transformation changing electoral allies. My point is that these new alignments were difficult to anticipate ex-ante.  They should also be visible. If opposition parties draw on their associational ties to mobilize voters, incumbents should be able to detect these attachments and understand their benefits. As such, incumbents have an interest in disrupting them. This assumption does not require that teachers’ unions be fully co-opted by parties and unable to occasionally act independently. This would be too strict a condition, and indeed, teachers’ unions sometimes offer criticism of their partisan allies and threaten strikes against them. For example, in 2000 and 2004 the Centro dos Professores do estado do Rio Grande do Sul went on strike against the PT despite their close electoral alliance [@goldfrank_restraining_2003]. I relax this assumption so that a perceived close connection between the opposition and teachers is sufficient to provide a substantial incentive to decentralize. 

Whether or not decentralization actually weakened the opposition is difficult to observe both from the standpoint of incumbents and academic research. The policy feedback process is complex and always generates intended and unintended effects. My point is that politicians have ideas about how certain policies might advantage them down the road and that identifynig whether or not those incentives exist can get a long way to revealing the political logic of decentralization.

# Case Study Evidence

To draw the logic of this theory, I look at the case of El Salvador, described briefly at the opening of this paper, with reference to the contrasting case of Paraguay. Case selection follows a most-similar-systems design: despite their structural and political similarities, these countries differ on outcomes (see table 1). El Salvador pursued a broad-scale decentralization program that deeply transformed patterns of authority across public education, whereas Paraguay engaged in various education sector reforms that never truly decentralized education. By holding constant structural characteristics commonly used to explain why countries engaged in decentralization, I show that these are not sufficient to explain patterns of education governance reform. The argument developed above can explain this puzzle.

One notable difference is the fact that El Salvador was emerging from a civil war, while Paraguay was not. I argue this is not a problem for the paired comparison because in both countries social and political actors wished to turn the page on a violent past in rural areas. In El Salvador, my interviews with political leaders confirmed that both incumbents and revolutionary leaders were eager to end fighting and were optimistic of their chances to influence politics through elections (Interview ES7, ES14). In Paraguay, the Stroessner and Rodríguez regimes were notable for conflicts in the abandoned rural areas. The transition process, however, also generated optimism among political actors that violence was behind them [@nagel_unleashing_1999]. In neither case was the "return" of abandoned rural territories into the national fold in question. But it did present an opportunity to rethink education provision. More importantly, for my purposes, previously outlawed or weakened opposition groups were perceived as being on the cusp of obtaining political power. While difference remain, I view the transition process of both countries, occurring in a context of optimism about electoral competition, as analytically equivalent. 

\input{tables/table1.tex}

This analysis draws on four weeks of fieldwork in El Salvador, including interviews and reviews of historical material, two trips to the World Bank archives in Washington, DC, to assess documents that contained the history of EDUCO in the context of World Bank loans, and secondary sources. An interview methods section can be consulted in section A, page 3 of the appendix. The case study covers events that happened nearly 30 years ago, placing straightforward limitations on the data—key actors may no longer be alive, or they may adjust their recollections of events to fit dominant narratives. To overcome these issues, wherever possible, I triangulate interview data with archival data and secondary materials. The case of Paraguay is constructed mostly from secondary sources and a review of primary documents.

## El Salvador: Education Decentralization as Political Weapon

From 1940 to 1992 the provision of public education in El Salvador was strongly centralized, with the Ministry of Public Education (hereafter, the Ministry) controlling the establishment, funding, and administration of public schools [@aguilar_aviles_vistazo_1998;@segovia_flores_historia_2012]. The Salvadoran civil war (1979–1992), however, severely affected the provision of public education in rural areas because government forces targeted teachers, who they assumed to be supportive of the FMLN.^[Interviews ES1, ES3, ES6] As a former guerrilla member puts it: “The guerrilla [forces] destroyed the economic infrastructure while the [government] army destroyed the education infrastructure; that was the division of labor.”^[Interview ES14] Moreover, reports by the @world_bank_salvador:_1990 and @unesco_servicios_1990 pointed to low completion rates and high repetition rates in public primary education in a context of falling central government investment in education. Thus, the Salvadoran government faced two typical problems for developing countries: improving the quality and coverage of education in a context of falling revenue.

Changing political circumstances opened a window for education reform. In 1989 Salvadorans elected the (right-wing) ARENA leader Alfredo Cristiani with a mandate to end the ongoing civil war. In anticipation of the signing of the Peace Accords (in 1992), government leaders were eager to return a sense of normalcy to daily life. This included returning official education services to rural areas and improving the quality of services.^[Interview ES14] A critical goal of any education reform was to “gradually decentralize management responsibilities to the regional and local levels” [@world_bank_staff_1993 pp.21] enshrined in the administration’s Socioeconomic Development Plan for the years 1990–1994. The idea was that decentralization, by virtue of delegating authority to lower levels of government, would create efficiencies in the provision of education at the same time that it improved the quality of education since parents would have a bigger say in the administration of schools and supervision of teachers. In sum, decentralization efforts were couched in technical considerations.

Government efforts to decentralize education were spearheaded by the Minister of Education at the time, Cecilia Gallardo, a hardline conservative with strongly held neoliberal views [@edwards_jr_trajectory_2017]. Gallardo’s menu of policy options was constrained by her ideology and dominant ideas in international education policy circles, both favoring some sort of decentralization strategy. More pressing, however, were the emerging political realities. During the civil war, local communities organized locally run schools known as *escuelas populares* to restore their access to education in places abandoned by the government. These *escuelas populares* served the guerrillas' ideological and political objectives. Ideologically, schools emphasized a state of war against injustice and inequality. Madre Rosa, a teacher, recounts: “I consider myself a *guerrillera* [guerrilla fighter] because [by teaching] we are fighting against a system that oppresses the poor and has never allowed them access to school, not even to learn their ABCs” [@de_la_cruz_educacion_1983 pp.95]. With the civil war coming to an end, it became imperative to exclude leftist elements and their institutions from areas previously controlled by the guerrillas.

The political salience of education reform was highlighted by the upcoming elections. Under the terms of the peace agreement, the FMLN would be allowed to participate as a political party in free elections in exchange for dismantling its military structure [@de_soto_implementation_1995]. In the run-up to the 1994 elections, there were thus two major political parties: ARENA and the FMLN. ARENA believed the elections were going to be competitive and viewed the FMLN as a real political contender.^[The FMLN did not nominate its own presidential candidate, but it did join a coalition of parties to present a nominee. It did, however, run its own candidates for the legislature.]  As reported by @lehoucq_election_1995, “[u]ncertainty about the electoral strength of the Left unnerved many in government, the military and their conservative supporters” (181). These groups acted accordingly. ARENA-affiliated death squads implemented a low-level campaign of violence against FMLN leaders. Prior to the elections, six leaders of the FMLN political coalition were murdered by right-wing death squads. ARENA also took to the airwaves, warning voters that an FMLN victory would threaten the peace and imply a “return to the past” [@stahler-sholk_salvadors_1994]. In addition, there were widespread claims of irregularities in the election process [@lehoucq_election_1995].

This political context affected the strategies pursued by the government to reform public education. Extending public schooling to rural areas could have been achieved by using the existing education governance system, but education leaders thought it would be too slow.^[Interview ES11] Most importantly, public school teachers and their organizations were viewed with suspicion. Most public school teachers belonged to the *Asociación Nacional de Educadores Salvadoreños, 21 de junio* (ANDES), teachers' union. The union was viewed as a tool of the radical left with deep partisan attachment to the opposition.  To some, the history of ANDES and the FMLN is one and the same. In 1970 the Secretary General of ANDES, Mélida Anaya Montes, co-founded one of the five component groups of the FMLN. Thus, in my interviews it was common for individuals to lump both organizations together. As a prominent right-wing politician told me, “The guerrilla is *born* in the teachers’ unions. The first guerrilla fighters of El Salvador are teachers”.^[Interview ES10]  While most ANDES-affiliated teachers were not directly involved in the conflict, there is some evidence that their organization provided support to the FMLN through union dues. In a cable from 1986, the CIA reports that Miguel Castellanos, a high-ranking FMLN defecting officer, stated that for all intents and purposes FMLN controlled ANDES, and that about half of all economic support it received was funneled to the guerrillas.^[``El Salvadors’ Insurgents: Resurrecting an Urban Political Strategy” Copy 181, from September 1986 held in the CIA Archives. Sanitized Copy Approved for Release 2011/12/30 accessed on January 21, 2019 available: [here][link1].]

[link1]: https://www.cia.gov/library/readingroom/docs/CIA-RDP04T00794R000100260001-0.pdf

The organizational closeness between the FMLN and ANDES makes clear that the latter's mobilizational advantages were key to the opposition's electoral strategy. @sprenkels_after_2018, who entered Chalatenango in 1992, recalls: "All the communities participated in party organizing, practically down to the last family." In places where the guerrillas did not have a visible presence, they exerted control through other agents; for example, “the ANDES teachers' union had departmental (subnational) organizational networks in Chalatenango within which the FPL  maintained strong contacts” (p.88). In the run-up to the 1994 elections, teachers formed a large part of the FMLN's party infrastructure. Most importantly, education leaders were painfully aware of this union-opposition bond and sought to disrupt it. In a 1991 interview Cecilia Gallardo, then Minister of Education, complained about the unions’ political activity: “What ANDES should do is put on the FMLN’s t-shirt, to which they have always belonged by the way, and go campaigning,”^[From a newspaper article in 1991. “Una crisis de libros y lectores: Licenciada de Cano," La Prensa Gráfica (February 11, 1994).]  implying that the teachers’ union was merely an extension of the FMLN, and that their organizational advantages were used not only to secure the interests of their affiliates but also to support their partisan allies.

The dilemma facing the government, then, was how to place the *escuelas populares* and newly built schools under state control without granting the FMLN a further electoral advantage in rural areas by placing their allies and brokers—ANDES—in these communities. ARENA and Cecilia Gallardo found balance in EDUCO, a decentralization initiative under which schools were managed by elected parent councils,^[Acronym for *Asociaciones Comunales para la Educación* or ACEs.] but the central government maintained authority over naming principals, funding schools, and setting the curriculum.

Two features of the program reveal its political logic. First, EDUCO was most heavily targeted toward rural areas, and specifically, FMLN strongholds. As noted by a politician: “The EDUCO model served as a way to transition the guerrilla territory [from the FMLN to the government].”^[Interview ES11] In the period between 1991 and 1993, EDUCO quadrupled in size by targeting communities precisely in FMLN-controlled areas. On the eve of the 1994 elections, EDUCO covered roughly 2,300 teachers and 74,000 students (about 17% of all primary school children), almost all of them in rural areas [@ornelas_education_2019]. However, the government subsequently expanded EDUCO to urban areas, eventually covering 40% of all children (CITE). 

Second, EDUCO was built to target teacher union organization. Under EDUCO teachers were hired on yearly contracts and were initially paid less than teachers in the official system.^[Eventually the World Bank, seeing EDUCO teachers jump ship whenever they could, advocated for a raise in the EDUCO pay schedule. This was after the 1994 elections. One of the unanticipated ramifications of the creation of parallel teaching tracks was that every EDUCO teacher wanted to enter the non-contracting option (Interviews ES1, ES6, ES12).] Teaching in EDUCO schools was therefore not appealing to tenured teachers affiliated with ANDES.^[Interviews ES5, ES6, ES12] At the same time, EDUCO teachers were required to meet a minimum set of teaching qualifications—i.e., obtaining some sort of degree. This had the effect of excluding instructors who had been teachers in *escuelas populares*.^[To allay the concerns of parents who had warmed up to teachers in *escuelas populares*, the Ministry of Education offered these teachers a path to formalization (Interview ES12). Instructors who had taught in *escuelas populares* would be eligible to teach under EDUCO if they underwent special trainings (Interview ES11). In practice, however, this had the effect of limiting the number of teachers in *escuelas populares* who would become EDUCO teachers, since it entailed leaving their posts, unpaid, for a significant amount of time. While it might be the case that 80 to 90% of teachers accepted the Ministry’s offer, as suggested by a former Ministry official (Interview ES11), the immediate job openings clearly presented a political opportunity for ARENA.] Finally, available teaching jobs were offered to ARENA sympathizers. In theory, the hiring of teachers was run through the parent councils. In practice, however, ACEs would often rubber-stamp the decisions made by government-appointed principals.^[Interviews ES4, ES7, ES8] The relationship between school principals and rural parents (some unable to read) was unequal, and principals thus maintained substantial authority in their communities. In one of my interviews a former EDUCO teacher admitted that he had been hired on the recommendation of a local ARENA party leader who had contacted the school principal.^[Interviews ES9] While it is impossible to know how common this practice was, this anecdote showcases that at different levels of government, the party’s commitment to EDUCO as a tool to improve educational outcomes was limited.

A final question remains: why would ANDES and the FMLN idly stand by as their political mobilization apparatus was dismantled? ANDES and its partisan allies were unusually weak in the aftermath of transition for two reasons. First, Gallardo was clever in designing the reform, both by developing a program (EDUCO) with roots in rebel organizations and by leveraging the support of the international community and prominent local actors.^[Interviews ES1, ES2, ES3] The fact that EDUCO schools were based on *escuelas populares* was meant to signal the program as a compromise with left-wing actors, despite the fact that it fully gutted *escuelas populares* and replaced them with a scheme favorable to the government. Moreover, the support of international organizations, including the World Bank, on a topic that all agreed was important—the improvement of education—placed ANDES on the defensive.

Second, most political actors viewed the transition period as tenuous, and therefore acted cautiously. Interviewees on the left and the right confirmed that the general sense among the population was one of exhaustion with social conflict.^[Interviews ES11, ES14] Thus, while ANDES mobilized large protests on a few occasions against education decentralization reforms, including a more severe decentralization program modeled after the Chilean experience,^[Interviews ES1, ES3] it eventually found itself bargaining with the government over the program. What did ANDES leaders extract from the government? The evidence is unclear, but some of my interviewees suggested that Gallardo promised ANDES leaders that most urban teachers (their electoral base) would remain excluded from the EDUCO system.^[Interviews ES5, ES11] Division within the FMLN was also significant. Some officers believed that an FMLN win would reverse democratization, splintering the organization.^[Interviews ES14] Thus, the FMLN was in a weak position to defend ANDES. In short, the specific dynamics of the transition and post-conflict period meant two things: 1) incumbents were relentless in their efforts to decentralize education while 2) ANDES and the FMLN were in no position to fully prevent it. The result was education governance decentralization.

The Salvadoran case study shows the political-electoral origins of education reform. The presence of an associational tie between teachers and the opposition incentivized ARENA to advance a very specific decentralization program meant to weaken the organizational capabilities of teachers, and therefore the opposition. EDUCO cannot be explained simply as a strategy to expand public education to war-torn regions; the logic of its expansion and its operation in practice reveals a very carefully targeted program. An unresolved question is whether education leaders, aside from their political motivations, actually believed that the program would improve the coverage and quality of education. My argument is entirely consistent with this possibility: even if they thought EDUCO would improve education outcomes, it is improbable that they would not have seen an electoral benefit to the excluding of ANDES-affiliated teachers from the countryside. Indeed, in the absence of such an incentive, reform would not have been advanced.

## Paraguay: No Decentralization Proposals in the Absence of Union-Opposition Ties

The history of public education governance in Paraguay is like that of El Salvador. Starting in 1933 the national government made a concerted effort to control education directly, generating a fairly centralized education system which remained untouched for most of the century. By the end of the 1980s this education system had delivered poor results, especially in comparison with other Latin American countries. Reports commissioned by the @world_bank_paraguay_2002 pointed to low completion and achievement levels among students, particularly those from rural communities and from poor backgrounds. In 1990 primary completion rates sat at 65%, roughly on par with El Salvador, and were lower still for the rural student population which accounted for half of all students [@world_bank_world_2018]. In sum, there was a technical case to be made for the improvement of public education. 

The overarching concern of the conservative party, the *Colorados*, however, was to remain in power under democracy, not education reform. General Stroessner was toppled in 1989 by his chosen successor General Andrés Rodríguez, who subsequently initiated a transition to democracy. Democracy brought new challenges for incumbents. The first sign of trouble emerged in 1991 when Carlos Filizzola of the *Asunción Para Todos* party was elected mayor^[Technically, Filizzola was elected *intendente municipal*.] of Asunción despite trailing the *Colorado* candidate Juan Manuel Morales in official polling.^["Paraguay celebra sus primeras elecciones municipales directas." El País, 26 of May, 1991, accessed on March 15, 2020 available: [here][link2].] Filizzola's victory offered a warning: in the post-transition period, the Colorados could not assume they would triumph at the ballot box [@velazquez_seiferheld_relaciones_2018]. In the run-up to 1993 this was confirmed by official polling, which placed the *Colorado* candidate, Juan Carlos Wasmosy, in an embarrassing third place. 

[link2]: https://elpais.com/diario/1991/05/27/internacional/675295206_850215.html

Counter to the expectations of the technical and ideational arguments, education reform under the right-wing *Colorados* was slow-moving. Recognizing that *something* had to be done, incumbents took a hands-off approach. In 1990 President Rodríguez established an education reform council, *Consejo Asesor de la Reforma Educativa* (CARE), and charged it with charting the path forward on education. To the relief of reformers, council members were well-respected independent experts who exhibited relative autonomy from the government in their deliberations [@nunez_reforma_2002]. CARE considered and debates many different types of education reforms, influenced in part by prominent ideas among reform advocates, international organization, and the Harvard Institute for International Development [@rodolfo_alisis_2014]. CARE considered decentralization a pillar of education reform [@rivarola_reforma_2000]. Indeed, CARE's first proposal, in 1992, centered reform around three objectives: 1) administrative changes in the Ministry of Education (MEC), 2) the selection of teachers, and 3) decentralization.^["Reforma Educativa, Compromisode Todos" 1992.] 

Incumbents, however, never seriously entertained decentralization. The reason is that teachers were not allied with the opposition; in fact, for most teachers the reverse was true. During the Stroessner era, public sector jobs were tightly controlled by the dictatorship. To be eligible for a teaching position, for instance, individuals  needed to be *Colorado* party members and required a written recommendation from a mid-tier party or military leader [@rivarola_reforma_2000]. The largest teachers' union, the *Federación de Educadores del Paraguay* (FEP), with roots in dictatorship, was thus aligned with the *Colorados* [@lopez_sindicatos_2009; @becker_sindicatos_2009]. The practice of appointing teachers on the basis of their political affiliation continued well into the 2000, when the *Colorados* were voted out of office in 2008 (with the election of Fernando Lugo). Institutional constraints, including tenure protections made dismissals of civil servant posts unheard of and thus extremely valuable [@schuster_when_2016].

Predictably, FEP mobilization was an important part of the *Colorado* electoral strategy. In the run-up to the 1993 elections, teachers were mobilized in rural areas to get out the vote in favor of the *Colorados*, leading to a high-profile clash with the Minister of Education, Raúl Sapena Brugada, who resigned rather than oversee a partisan education system [@nickson_wasmosy_1997; @harding_paraguay_1993]. Despite partisan alignments, FEP and other unions depended on centralized education governance to obtain benefits for their membership. FEP EXAMPLE

From political perspective, the *Colorados* might yet reap an electoral reward from decentralization by weakening the teacher labor movement more generally. Democratization meant that new unions could form and participate in politics. Other, smaller teachers' unions active around this time included the *Organización de Trabajadores de la Educación del Paraguay-Sindicato Nacional* (OTEP) and the *Unión Nacional de Educadores* (UNE) [@becker_sindicatos_2009]. These organizations had a contentious relationship with the incumbents, deploying strikes to bargain over the improvement of teacher working conditions. OTEP-led protests in 1990 and 1993, culminated in significant wins for teachers, including a flat wage increase of 30% [@nunez_reforma_2002]. If education decentralization hurts the organization of teachers' unions, pursuing decentralization as CARE suggested could have weakened unions ability to mobilize for their interest. However, given the importance of FEP to the *Colorados* this was a nonstarter. 

Efforts to decentralize education therefore fizzled, despite a larger push to grant subnational units (*departamentos* and *municipios*) political and fiscal authority in the 1992 constitution. For example, the 1993 *Ley Orgánica Municipal* provided that municipalities *could* run their own schools if they so wished, but apportioned no national funds for such activity. Indeed, with the exceptions of Asunción y San Lorenzo there are no municipal schools in Paraguay [@gaete_estudio_2012]. At the department level, the 1998 *Ley General de Educación* mandated that the MEC establish subnational council to coordinate efforts with the *departamentos*. But these councils are at best weak, advisory bodies to the MEC; they do not hire teachers, set education policy, nor do they administer or manage schools [@letelier_s_education_2018]. 

In sum, by the end of the 1990s education in Paraguay can only be characterized as strongly centralized, if with some allowances for subnational units to convey concerns over education to national authorities. Looking back on the promise of the education reform movement, the perception among many reform advocates is that things did not change radically. In the evaluation of a teacher not affiliated with FEP: "they have engaged in curriculum reforms, that theoretically and pedagogically where OK...but everything pretty much remains as before" [@palacios_de_asta_descentralizacion_2002]. This uneven commitment to decentralization by the *Colorados*--who remained in power until 2008, and thus had plenty of further opportunities to pursue reform--can only be explained by the absence of a opposition-union link. 

# Implications for Existing Work and Future Research Avenues 

As to Nicaragua more specifically, a review of the secondary literature on education decentralization in that country suggests that my argument could explain its decentralization program as well. First, the decentralization program was initially couched in technical terms: the expansion of and increased efficiency in public education (Arcia and Belli 1990). Second, the government advanced the project in the absence of concurrent political or fiscal devolution; as such it looks a lot like EDUCO (Gershberg and Meade 2005). Third, as reported by Kubal (2003), the program was pushed by the incumbent, conservative government (Chamorro’s) and opposed by the Sandinista-aligned teacher’s union. There are other intriguing parallels and a few key differences but I am afraid I have not done the work to properly understand the politics around it. For example, the fact that Nicaragua features a more fragmented teacher union environment might either strengthen my argument or challenge it. I have left this task to future work but I have made a nod to this end in the conclusion. 

That politics structures education systems is an increasingly important point in the small literature of education [@ansell_political_2013; @kosack_education_2012; @moe_comparative_2016]. My work moves this literature forward by making explicit that changes to education systems can serve as an electoral strategy to weaken political opponents. My work also asks us to reassess conventional wisdom regarding the link between electoral competition and the provision of public services. Where other authors have highlighted how electoral competition can lead to an expansion of social policy  [@dion_international_2008;@garay_social_2016;@de_la_o_crafting_2015], I show that electoral competition can lead incumbent political parties to make targeted changes to social policy to hurt the preexisting organizational and mobilizational advantages of their rivals. 

My focus on relatively understudied Latin American countries reveals something important about the politics of decentralization. The literature on decentralization conceptualizes education reforms of the type considered here as administrative decentralization. This view of decentralization works well in larger countries with a history of subnational units articulating demands from central government authorities, and in assessing the final outcomes of decentralization in multilevel countries. For example, in Colombia and Argentina, education decentralization was part of a larger decentralization package that was supposed to devolve more authority to subnational units of government. But this frame is an awkward fit for smaller, unitary countries -- like El Salvador and Paraguay, but also Guatemala, Honduras, and Nicaragua. Looking for an explanation for education decentralization that generalizes to these and other countries requires thinking about the education sector specifically and it's connection to politics. I thus focus on evaluating answers provided by the older literature on education reform in Latin America. 

The biggest theoretical takeaway from this dissertation is that electoral incentives matter in assessing why the implementation of education decentralization succeeds. In competitive electoral systems, governments view policies through the lens of reelection. Thus, even projects seemingly far-removed from national politics, like education decentralization, have electoral origins and consequences. In the case of education, incumbents will do whatever they can to hamstring the organization of teachers if this improves their own electoral odds.



```{r setupfig3, include=FALSE}
## Bayesian Regressions

## Load Packages
library(tidyverse)
library(socviz)
library(hrbrthemes)
library(myriad)

## Read in data
pol<-read.csv("data/pres_political_data.csv") %>%
  mutate(year = as.factor(election_year),
         rates = completion_rate/10,
         oul = teachers_with_opp)

mini<-pol %>%
  select(education_decentralization_reform, completion_rate,
                        right_center_of_gravity, imf_education_loans, teachers_with_opp, 
                        polity, oul)
#stargazer(mini) # For appendix

data<-read.csv("data/presidential_admin.csv")
fig4<-ggplot(data, aes(x=election_year, y=decentralization_level,
                       linetype=as.factor(teachers_with_opp),
                       color= as.factor(teachers_with_opp),
                       shape=as.factor(teachers_with_opp))) +
  geom_jitter(size=2) +
  geom_smooth(method = lm, se= F, size=3) +
  labs(x="Election Year", y="Decentralization Level", linetype="Opp-Teach", color="Opp-Teach") +
              theme_bw()+
            theme(panel.border = element_rect(fill = NA, size = 1),
                  legend.position = "none") +
              coord_cartesian(xlim=c(1980,2010), ylim = c(0,1))+
            labs(y="Level",
                 x="Year",
                 title = "Education Decentralization",
                 subtitle = "Across Opposition-Union Links") 

```

```{r, plotfig3, echo=FALSE, message=FALSE, warning=FALSE, error=FALSE, fig.cap = "Association Between Opposition-Union Links and Decentralization", fig.width=4, fig.height=3, cache= F, label=fig4}
fig4
```

Does this travel to other contexts? The plot contains two types of presidential administration observations: those in which there is a union-opposition link (OUL) and those without. Those with OUL are represented by green triangles, those without OUL are represented by red dots. I have also overlaid regression lines. The trend for administrations with OUL is in dashed green, the trend for administration without OUL are in solid red. 


My analysis offers important public policy lessons. A robust literature on education reform emphasizes the role that teachers’ unions play in blocking reforms [@bonilla_pensiones_1998; @corrales_politics_1999; @graham_improving_1999; @moe_comparative_2016; @diaz_rios_when_2016]. Yet the argument developed here asks us to reassess the commitments of elected leaders to improve education above all else. The simple point that incumbents want to win elections has implications for how we understand why education reforms are successful. We should expect, for instance, that decentralization efforts are least likely to be pursued in good faith whenever unions are close with the opposition, because the overriding concern of parties should be to weaken the opposition. Such reforms, however, are positioned to fail, both because they will never get the buy-in of teachers and because a return to power by the opposition will dismantle them (such as in the case of El Salvador). Conversely, changes to the education system should be most likely to stick when teachers are unattached to political parties because incumbents are motivated to reform for reasons other than weakening opponents, leading them to engage unions extensively to do so.  

Perhaps, then, countries decentralized because the electorate demanded it. Latinobarómetro data from 1990–1995, however, does not support this view. While there is no data on support for decentralization projects, there is data on support for education systems generally. When asked whether the quality of education has decreased or increased over the past five years a plurality of respondents say that education quality has increased by a lot. This echoes more recent findings suggesting that, despite the oft-cited gaps in coverage and quality, Latin Americans have long been more satisfied with education than their counterparts in other regions [@pages_beyond_2008], and that generally the electorate does not reward elites for investment in quality-improving reforms [@boas_competence_2021]. 

\theendnotes
